---
title: "Numbers and weight by haul and length classes"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{Numbers and weight by haul and length classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

The goal of this vignette is to demonstrate how to derive numbers and weight by
haul and length classes based on the DATRAS database. This requires following
steps:

**Outline:**

1. Create length spectrum
2. Create weight spectrum
3. Aggregated length classes
4. Summary
5. References


The vignette uses the the *DATRASextra* package:

```{r}
library(DATRASextra)
```

and assumes that a cleaned *DATRAS* database is available (see tutorial
vignette) for information on how to get a cleaned *DATRAS* file. Here, we use
the

```{r}
data("dab")
```


## *Create length spectrum*

```{r}
lpars <- checkLength(dab)
```



```{r}
dab <- addSpectrum(dab)
```


Alternatively, custom length breaks with e.g. maximum length threshold:


```{r}
l.cut.off <- lpars$lPars$median * 5  ## find good cut-off
## or l.cut.off <- lpars$lPars$perc
cm.breaks <- attr(lpars$N,"cm.breaks")
nbyl <- colSums(lpars$N)[which(cm.breaks <= l.cut.off)]
threshold <- cm.breaks[max(which(nbyl != 0))]
```



```{r}
## add length data with custom breaks
by <- DATRAS:::getAccuracyCM(dab)
cm.breaks <- seq(lpars$lPars$min, threshold + by, by = by)
LngtCode2cm <- c(. = 0.1, `0` = 0.5, `1` = 1, `2` = 2, `5` = 5)
while (length(cm.breaks) > 100 && which(by == LngtCode2cm) != length(LngtCode2cm)) {
    by <- LngtCode2cm[which(by == LngtCode2cm) + 1]
    cm.breaks <- seq(lpars$lPars$min, threshold + by, by = by)
}

dab <- addSpectrum(dab, cm.breaks = cm.breaks, by = by)
```




## *Create weight spectrum*


```{r}
wpars <- checkWeight(dab)
```


```{r}
dab <- addWeight(dab, maxL = threshold)
```


Alteratively, ...

```{r}
# dab <- addWeightEmpirical(dab)
```




## *Aggregated length classes*

If for example only 2 or 3 length classes: ....

```{r}
cuts <- c(0, 10, 20, 30, Inf)
by <- diff(attr(dab, "cm.breaks"))
midL <- attr(dab, "cm.breaks")[-length(attr(dab, "cm.breaks"))] + by/2
ind <- as.integer(cut(midL, cuts, right = TRUE, include.lowest = TRUE))
nbins <- length(cuts) - 1
nNew <- wNew <- matrix(0, nrow = nrow(dab$N), ncol = nbins)
for (j in seq_len(nbins)) {
    cols.in.bin <- which(ind == j)
    if (length(cols.in.bin) > 0) {
        nNew[, j] <- rowSums(dab$N[, cols.in.bin, drop = FALSE])
        wNew[, j] <- rowSums(dab$Wgt[, cols.in.bin, drop = FALSE])
    }
}
colnames(nNew) <- colnames(wNew) <- paste0("(",cuts[-length(cuts)], "-", cuts[-1],"]")

dab$HaulN <- nNew
dab$HaulWgt <- wNew
```


Plot:


```{r}
par(mfrow = n2mfrow(ncol(nNew), asp = 2),
    mar = c(3,3,2,2), oma = c(2,2,0,0))
for (i in 1:ncol(nNew)) {
    plot(dab[[2]]$lon, dab[[2]]$lat,
         ty = "n",
         xlab = "", ylab = "",
         main = colnames(dab[[2]]$HaulN)[i])
    ind <- which(dab[[2]]$HaulN[,i] > 0)
    points(dab[[2]]$lon[ind], dab[[2]]$lat[ind],
           cex = dab[[2]]$HaulN[,i][ind] /
               max(dab[[2]]$HaulN[,i][ind]) * 3,
           col = i)
}
mtext("Longitude", 1, outer = TRUE)
mtext("Latitude", 2, outer = TRUE)
```




## *Summary*

This vignette introduced the main features and workflow of the *DATRASextra*
package for ...



## *References*
