---
title: "Data processing and quality control"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Data processing and quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = TRUE,
                      eval = TRUE,
                      error = FALSE,
                      message = TRUE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width = 8,
                      fig.height = 7,
                      dev = "png",
                      dev.args = list(bg = "white"))

knitr::knit_hooks$set(
  plot = function(x, options) {
    paste0(
      "<div style='background:white; padding:0.5rem; display:inline-block;'>",
      knitr::hook_plot_html(x, options),
      "</div>"
    )
  }
)

oldpar <- par(no.readonly = TRUE)
par(bg = "white")
```

```{r, message=FALSE, warning=FALSE}
library(DATRASextra)
```

Every DATRAS analysis rests on a chain of processing decisions: which hauls
count as valid, how sub-sampled catches are raised, what a missing value looks
like, and which records are dropped along the way. Most of those decisions are
made before a user writes a single line of analysis code.

This vignette documents that chain end to end. For each step it states what is
done, whether it happens automatically or on request, how to inspect its effect,
and how to change or switch it off. It also lists checks that **DATRASextra**
deliberately leaves to the user, because the right answer depends on the
analysis.

Two principles guide the design:

- **Nothing is silently discarded.** The mandatory filters in `clean_datras()`
  are the DATRAS-recommended minimum, and the function prints what will be
  removed *before* removing it.
- **Checks report, they do not delete.** `check_outliers()`, `check_lengths()`
  and `check_weights()` are opt-in, and `check_outliers()` returns the object
  unchanged unless explicitly asked to remove hauls.


# Overview of the pipeline

| Stage                | Function                                                                                                     | What happens                                                                                                                          | Automatic?                         |
|----------------------|--------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------------|------------------------------------|
| 1. Download          | `download_datras()` (via `DATRAS::getDatrasExchange()`)                                                      | Query ICES web service, harmonise column names, convert the `-9` sentinel to `NA`, match orphan `CA` records, archive as exchange zip | automatic                          |
| 2. Read              | `read_datras()` (via `DATRAS::readICES()`)                                                                   | Parse exchange CSV, drop truncated files, match orphan `CA` records, drop duplicated hauls, combine survey-years                      | automatic                          |
| 3. Derived variables | `DATRAS::addExtraVariables()`                                                                                | Create `haul.id`, `LngtCm`, `Species`, `Count`, `lon`/`lat`, `abstime`, `Roundfish`; treat hauls without `HL` records as zero-catch   | automatic                          |
| 4. Cleaning          | `clean_datras()`                                                                                             | Filter on `HaulVal` and `StdSpecRecCode`, optionally impute depth and harmonise species, optionally subset                            | automatic filters, optional extras |
| 5. Checks            | `check_outliers()`, `check_lengths()`, `check_weights()`, `add_swept_area()`                                 | Flag implausible values, summarise length and weight distributions, report swept-area missingness                                     | opt-in                             |
| 6. Inspection        | `print()`, `summary()`, `plot_datras_overview()`, `plot_length_distribution()`, `plot_species_composition()` | Tabular and visual review                                                                                                             | opt-in                             |


Throughout we use the bundled example data set `mini`: four surveys, seven
gears, five species, 2022-2023. Unlike the `dab` example, `mini` has not been
pre-cleaned, so every filter described below actually does something.

```{r}
mini
```


# Stage 1: Download

`download_datras()` wraps `DATRAS::getDatrasExchange()`, which in turn queries
the ICES DATRAS web service. Several things happen before anything reaches
disk.

```{r, eval = FALSE}
## One survey, selected years, archived under <path>/NS-IBTS/
x <- download_datras(path = "data/datras", surveys = "NS-IBTS", years = 2020:2023)
```

## Which surveys and years are requested

Available surveys, years and quarters are read from the ICES web service. If the
server does not answer within `timeout` seconds (default 10), the function falls
back to a cached survey list and, for years, to whatever zip files are already
present locally. This keeps a partially completed download resumable when the
connection is unreliable, but it also means that an offline session may silently
work from a stale survey list.

Two surveys are skipped by default because they are test entries or do not
contain the full set of required tables: `Test-DATRAS` and `NS-IBTS_UNIFtest`.
Set `include_flagged = TRUE` to download them anyway.

The `download_hl` and `download_ca` arguments control whether the
length-frequency and biological tables are fetched at all. Omitting whichever
table an analysis does not need reduces both download time and memory use; `CA`
is the usual candidate, since many analyses need only catch-at-length.

## Harmonisation applied on download

`DATRAS::getDatrasExchange()` applies the following, in order:

1. **Column renaming** (`DATRAS::renameDATRAS()`). Names are capitalised, and
   the web-service names are mapped onto the exchange-format names:
   `ValidAphiaID` becomes `Valid_Aphia`, `LngtClass` becomes `LngtClas`,
   `AgeRings` becomes `Age`, and `CANoAtLngt` / `NoAtLngt` become `NoAtALK`.
   This is why the same variable has different names in the ICES documentation
   and in an R object.
2. **Missing-value conversion** (`DATRAS::minus9toNA()`). DATRAS encodes missing
   numeric values as `-9`. Every numeric column is scanned and `-9` is replaced
   by `NA`. Note that this applies to numeric columns only: a `-9` stored as
   text is left untouched.
3. **`StatRec` back-fill.** If the `CA` table has no `StatRec` column, it is
   filled from `AreaCode`.
4. **Derived variables** are added (see Stage 3 below).
5. **Orphan `CA` records** are matched back to hauls with
   `DATRAS::fixMissingHaulIds()` using `strict = TRUE`, so ambiguous records are
   left unmatched rather than attributed to an arbitrary haul.
6. **Character columns are converted to factors.**

## What gets archived

Before writing, `download_datras()` strips the derived columns again and maps
the names back to the exchange-format spelling, so the archived zip is a
faithful DATRAS exchange file rather than an R-specific artefact. Anyone can
re-read it, with **DATRASextra** or with any other tool that understands the
exchange format. Files are named `<survey>_<year>.zip` inside a per-survey
subdirectory.

This separation matters for reproducibility: the archive is the raw record, and
every processing step described in the rest of this vignette is reapplied from
it each time the data are read.

```{r, eval = FALSE}
## Re-download and replace existing files
download_datras(path = "data/datras", surveys = "NS-IBTS", years = 2020,
                overwrite = TRUE)
```

Worth checking after a download: that a file exists for every year you expect,
and that none of them is suspiciously small. `read_datras()` performs the size
check for you (next section), but an unexpectedly missing year usually means the
survey was not conducted rather than that the download failed.


# Stage 2: Reading data into R

```{r, eval = FALSE}
x <- read_datras("data/datras", surveys = "NS-IBTS", years = 2020:2023)
```

## Parsing and file-level checks

A DATRAS exchange CSV contains three stacked blocks, each with its own header
line. `DATRAS::readICES()` locates the header lines, reads each block
separately, and applies `na.strings = c("-9", "-9.0", "-9.00", "-9.0000")` so
that the missing-value sentinel is converted on the way in. It then verifies
that the number of data rows plus header lines equals the number of lines in the
file, and aborts with `"csv file appears to be corrupt"` if it does not.

`read_datras()` adds a guard of its own: files smaller than `min_file_size`
(default `1e4` bytes) are reported and excluded, because a truncated or empty
download would otherwise raise a confusing parse error. A year in which a survey
was not conducted typically produces such an empty file.

## Matching biological records to hauls: the `strict` argument

Records in `CA` frequently lack the station and haul numbers that make up
`haul.id`. `DATRAS::fixMissingHaulIds()` matches them back to a haul using
survey, year, quarter, country, ship and statistical rectangle. When that
combination identifies more than one candidate haul, the two settings diverge:

- `strict = FALSE` assigns the record to **one of the candidate hauls at
  random**.
- `strict = TRUE` leaves the record as `NA`, and it is dropped by any subsequent
  subsetting.

Which behaviour is right depends on the analysis: random assignment preserves
sample size for aggregate age-length keys, while leaving records unmatched is
the honest choice when individual records must be attributable to a specific
haul.

```{r, eval = FALSE}
## Attribute ambiguous CA records to an arbitrary haul
x <- read_datras("data/datras", strict = FALSE)
```


## Combining survey-year files

Files are read one at a time and combined. Two safeguards apply:

- **Duplicated haul identifiers.** The same haul appearing in more than one file
  is reported by `haul.id` and removed. Note that the two code paths behave
  slightly differently: the sequential reader (`ncores = 1`) keeps the first
  copy encountered and drops later ones, while the parallel reader removes all
  copies. If duplicates are reported, the underlying archive is worth
  investigating rather than accepting either behaviour.
- **Type-consistent binding.** `c.datras_raw()` guards against a column that is
  a factor in one year and an integer in another. Plain `rbind()` would use the
  factor's level vector as the type and reinterpret the other year's integers as
  factor codes, silently turning valid values into `NA`.

For large reads, `prune = TRUE`, `drop_hl = TRUE` and `drop_ca = TRUE` reduce
peak memory by trimming columns or dropping whole tables inside the reader, and
`surveys` / `years` restrict which files are opened at all.


# Stage 3: Variables created while reading

The columns most analyses actually use are not in the exchange file at all: they
are constructed by `DATRAS::addExtraVariables()` during both download and read.

## Haul identifier

```
haul.id = Survey:Year:Quarter:Country:Ship:Gear:StNo:HaulNo
```

This is the key that links `HH`, `HL` and `CA`.

```{r}
head(levels(mini[["HH"]]$haul.id), 3)
```

## Length in centimetres

`LngtClas` is reported in units that depend on `LngtCode`. The conversion to
`LngtCm` uses:

| `LngtCode` | Multiplier for `LngtClas` | Meaning                        |
|------------|---------------------------|--------------------------------|
| `.`        | 0.1                       | reported in mm                 |
| `0`        | 0.1                       | reported in mm, 0.5 cm classes |
| `1`        | 1                         | reported in cm, 1 cm classes   |
| `2`        | 1                         | reported in cm, 2 cm classes   |
| `5`        | 1                         | reported in cm, 5 cm classes   |

There is a trap here worth stating explicitly. A **second**, different table with
the same codes is used by `get_accuracy_cm()` (and `DATRAS::getAccuracyCM()`) to
describe the *bin width*: `.` = 0.1, `0` = 0.5, `1` = 1, `2` = 2, `5` = 5. The
first table converts a recorded value to centimetres; the second says how wide
the length class is. They are not interchangeable.

```{r}
## Which length codes occur, and what resolution do they imply
table(mini[["HL"]]$LngtCode, useNA = "ifany")
```

```{r}
## The coarsest resolution present is used for automatic length binning
get_accuracy_cm(mini)
```

Mixed codes are common when several surveys or countries are combined, and
`get_accuracy_cm()` warns and falls back to the coarsest resolution. Empty
`LngtCode` values are assumed to be 1 cm. If that assumption is wrong for your
data, pass `cm_breaks` or `by` explicitly to the length functions rather than
relying on the automatic choice.

## Species names

`Species` is resolved from `SpecCode` using either an ITIS/TSN lookup table or a
WoRMS lookup table, depending on whether `SpecCodeType` is `"W"`. Older survey
years often use TSN codes while recent years use WoRMS codes, so both paths can
occur in a single combined object.

```{r}
table(mini[["HL"]]$SpecCodeType, useNA = "ifany")
```

## Raised catch numbers

This is the most consequential derived variable:

```
Count = HLNoAtLngt * (HaulDur / 60 if DataType == "C", else 1) * SubFactor
```

Two corrections are folded into one column:

- **Sub-sampling.** `SubFactor` raises the measured sub-sample to the whole
  catch.
- **Standardisation of reporting units.** `DataType == "C"` means numbers were
  reported per hour, so they are converted back to per-haul using the haul
  duration. `DataType == "R"` (raw) and `"P"` are used as reported.

A combined object frequently mixes reporting conventions:

```{r}
table(mini[["HH"]]$DataType, useNA = "ifany")
```

Anything computed from `Count` therefore already accounts for sub-sampling and
per-hour reporting. Recomputing from `HLNoAtLngt` without applying both
corrections is a common source of error. Note also that `check_lengths()` warns
if `DataType == "S"` is present, since those hauls are treated as `"R"`.

## Time and position

- `lon` and `lat` are copies of `ShootLong` and `ShootLat`, that is the position
  where the gear was shot, **not** the midpoint of the tow. `HaulLat` and
  `HaulLong` hold the end position if a tow midpoint is needed.
- `abstime` is a decimal year, `timeOfYear` its fractional part, and
  `TimeShotHour` the time of day in decimal hours. These are convenient for
  smoothers over time or time of day.
- `Roundfish` is the North Sea roundfish area, derived from `StatRec`.

## Hauls without catch records

Any haul present in `HH` that has no rows in `HL` is interpreted as a
**zero-catch haul**. During reading, the upstream code prints a warning listing
the affected hauls. This is the correct treatment for a species-restricted
extract such as `mini`, where most hauls simply did not catch the five target
species, but it means a haul missing from `HL` for a data-handling reason is
indistinguishable from a genuine zero.

The split is reported by `print()`:

```{r}
mini
```

## Type coercions to be aware of

- `Year` and `Quarter` are converted to **factors**. `mini[["HH"]]$Year + 1`
  does not do what it looks like; use `as.integer(as.character(Year))`.
- All character columns become factors.
- `haul.id` in `HL` and `CA` is re-levelled to the levels present in `HH`, so
  records that do not correspond to a haul in `HH` become `NA` and are dropped
  by later subsetting.
- Duplicated rows in `HH` abort the read entirely with
  `"Duplicated rows found in HH data - data file is corrupt"`.


# Stage 4: Cleaning with `clean_datras()`

`clean_datras()` applies the standard DATRAS cleaning filters. With
`verbose = TRUE` (the default) it prints a full account of the data *before*
filtering, and `explain_codes = TRUE` adds the meaning of each code.

```{r}
surv <- clean_datras(mini, explain_codes = TRUE)
```

The report has three parts: the distribution of the three recording codes, the
missing-value counts for the variables that later steps depend on, and the
number of hauls removed by each filter.

## Filter 1: haul validity

Only hauls with `HaulVal` in `c("V", "N")` are retained.

| Code | Meaning                                                   | Kept |
|------|-----------------------------------------------------------|------|
| `V`  | Valid haul                                                | yes  |
| `N`  | No oxygen (BITS only)                                     | yes  |
| `S`  | Standard haul                                             | no   |
| `A`  | Additional valid station, not used for index calculations | no   |
| `C`  | Calibrated (BITS only)                                    | no   |
| `I`  | Invalid haul                                              | no   |
| `M`  | Pelagic midwater trawl (BITS only)                        | no   |
| `P`  | Partly valid haul (sensor problems)                       | no   |

`"N"` hauls are Baltic stations aborted because of oxygen depletion. They are
retained because the catch is valid for the duration fished, but they often have
an unusually short `HaulDur`, and `clean_datras()` says so when any are kept.
Whether to keep them is an analysis decision: they are legitimate observations,
but a swept-area index that does not account for the shorter tow will be biased.

## Filter 2: standard species recording

Only records with `StdSpecRecCode == 1` ("all standard species recorded") are
retained. Codes `2`, `3` and `4` mean that only pelagic, roundfish or individual
standard species were recorded, and `0` means none were. Keeping them would mix
hauls with different species coverage, which biases community-level and
multi-species analyses.

Note what is **not** filtered. `BySpecRecCode` is reported but never used as a
filter: bycatch recording protocols differ by survey, and which species are
usable under each code is an analysis-specific judgement. The relevant
per-code species lists exist in the source of `correct_species()` as reference
material but are deliberately inactive.

## Optional: imputing missing depths

Disabled by default. When enabled, missing `Depth` values are predicted from a
GAM of `log(Depth)` on a two-dimensional smooth of longitude and latitude, using
the hauls that do have a recorded depth. This requires the R package **mgcv**.

```{r}
surv_dep <- clean_datras(mini, impute_missing_depth = TRUE, verbose = FALSE)

## Missing depths before and after
c(before = sum(is.na(mini[["HH"]]$Depth)),
  after = sum(is.na(surv_dep[["HH"]]$Depth)))
```

The imputed values replace `NA` in place, and there is no flag distinguishing
them afterwards. If that distinction matters, record the affected haul
identifiers before cleaning:

```{r}
imputed_ids <- as.character(mini[["HH"]]$haul.id[is.na(mini[["HH"]]$Depth)])
imputed_ids
```

Spatial interpolation of depth is reasonable on a continental shelf and poor
across a shelf break. Inspect the result before relying on it.

## Optional: harmonising species

`correct_species = TRUE` (the default) calls `correct_species()`, which does two
things.

First, it collapses taxa that are not reliably identified to species in the
field down to genus level, updating `Species`, `Valid_Aphia` and `Rank`:
*Dipturus*, *Liparis*, *Chelon*, *Mustelus*, *Alosa*, *Argentina*,
*Callionymus*, *Ciliata*, *Gaidropsarus*, *Sebastes*, *Syngnatus*,
*Pomatoschistus* and *Gobius*. Records identified as *Nerophis ophidion*,
*Leusueurigobius*, *Neogobius* and *Argentinidae* are folded into the
corresponding genus.

Second, it corrects outdated or misspelled names: *Synaphobranchus kaupi*
becomes *Synaphobranchus kaupii*, and *Dipturus linteus* becomes
*Rajella lintea* with the current AphiaID.

The `Rank` column it adds makes the outcome visible:

```{r}
table(surv[["HL"]]$Rank, useNA = "ifany")
```

Set `correct_species = FALSE` to keep the original species assignments, for
example when reproducing an earlier analysis or when comparing against the raw
ICES extract.

## Optional: subsetting

`aphias`, `years`, `quarters` and `gears` are applied **after** the filters, as
a convenience:

```{r}
surv_sub <- clean_datras(mini, years = 2023, quarters = 1,
                         aphias = 127137, verbose = FALSE)
surv_sub
```

## Modifying or omitting steps

| Step                    | Argument                               | Default        | Effect of changing                     |
|-------------------------|----------------------------------------|----------------|----------------------------------------|
| Pre-cleaning report     | `verbose`                              | `TRUE`         | `FALSE` silences it, for scripts       |
| Code descriptions       | `explain_codes`                        | `FALSE`        | `TRUE` annotates each code             |
| Haul validity filter    | none                                   | always applied | reproduce manually with `subset()`     |
| Standard species filter | none                                   | always applied | reproduce manually with `subset()`     |
| Depth imputation        | `impute_missing_depth`                 | `FALSE`        | `TRUE` fills missing depths by GAM     |
| Species harmonisation   | `correct_species`                      | `TRUE`         | `FALSE` leaves names untouched         |
| Subsetting              | `aphias`, `years`, `quarters`, `gears` | `NULL`         | applied after filtering                |
| Alternative pipeline    | `do_fishglob`                          | `FALSE`        | `TRUE` delegates to `clean_fishglob()` |

The two mandatory filters have no on/off switch, but they are ordinary
`subset()` calls and are trivially replaced when a different rule is wanted:

```{r}
## Keep additional valid stations as well as valid and no-oxygen hauls
custom <- subset(mini, HaulVal %in% c("V", "N", "A"))

## Accept hauls where only individual standard species were recorded
custom <- subset(custom, StdSpecRecCode %in% c(1, 4))

## Then apply the optional steps only
custom <- correct_species(custom)
custom
```

Compare the haul count with `surv` above to see what the relaxed rule buys, and
decide whether the extra stations are usable for your purpose.

For the FishGlob workflow, `do_fishglob = TRUE` replaces the whole cleaning
stage with `clean_fishglob()`; see `vignette("articles/fishglob")`.


# Stage 5: Quality-control checks

These functions are deliberately not called by `clean_datras()`. They flag
values that are implausible rather than formally invalid, and whether a flagged
record should be excluded is a decision the analyst has to make.

## `check_outliers()`

Two independent mechanisms, both optional to act on.

### Rule-based checks

Fixed, physically motivated bounds. `strict = TRUE` is the default; `FALSE`
relaxes the upper bounds.

| Table | Variable                   | Rule (`strict = TRUE`)          | Rule (`strict = FALSE`) |
|-------|----------------------------|---------------------------------|-------------------------|
| `HH`  | `Quarter`                  | outside 1-4                     | same                    |
| `HH`  | `Year`                     | before 1965 or in the future    | same                    |
| `HH`  | `TimeShot`                 | not a valid HHMM time           | same                    |
| `HH`  | `HaulDur`                  | outside 0-240 min               | outside 0-360 min       |
| `HH`  | `ShootLat`                 | outside -90 to 90               | same                    |
| `HH`  | `ShootLong`                | outside -180 to 180             | same                    |
| `HH`  | `Depth`                    | outside 0-2000 m                | outside 0-3000 m        |
| `HH`  | `GroundSpeed`              | outside 0-6 knots               | outside 0-8 knots       |
| `HH`  | `DoorSpread`               | outside 0-250 m                 | outside 0-300 m         |
| `HH`  | `WingSpread`               | outside 0-80 m                  | outside 0-120 m         |
| `HH`  | `WingSpread`, `DoorSpread` | wing spread exceeds door spread | same                    |
| `HL`  | `LngtCm`                   | outside 0-200 cm                | outside 0-300 cm        |
| `CA`  | `Sex`                      | not in `""`, `F`, `M`, `U`      | same                    |
| `CA`  | `Age`                      | outside 0-50, or non-integer    | outside 0-80            |
| `CA`  | `NoAtALK`                  | negative or non-integer         | same                    |
| `CA`  | `IndWgt`                   | outside 0-1e5 g                 | outside 0-3e5 g         |
| `CA`  | `LngtCode`                 | not in `""`, `.`, `0`, `1`      | same                    |
| `CA`  | `LngtClas`                 | outside 0-2000, or non-integer  | outside 0-3000          |

Rule violations get severity `"invalid"`.

### Percentile checks

Enabled with `pct = TRUE`. Within groups, values outside the requested
percentiles are flagged with severity `"extreme"`. Defaults:

- `pct_probs = c(0.01, 0.99)`
- `pct_by`: `HH` grouped by survey, quarter, gear and ship; `HL` and `CA` by
  survey, quarter, gear and species
- `pct_vars`: `HaulDur`, `Depth`, `DoorSpread`, `WingSpread` in `HH`; `LngtCm`
  in `HL`; `Age`, `IndWgt`, `LngtClas` in `CA`
- `pct_min_n = 50`: groups with fewer observations are skipped rather than
  producing unstable thresholds
- `pct_log_vars`: `IndWgt` percentiles are computed on the log scale

Grouping matters. Comparing a beam trawl against a GOV trawl on the same
threshold produces nonsense; the defaults compare like with like.

### Running the checks

```{r}
surv <- check_outliers(surv, pct = TRUE)
```

By default nothing is removed: the object comes back unchanged, with four
attributes attached.

```{r}
## Every flagged record, one row each
report <- attr(surv, "outlier_report")
dim(report)
head(report[report$severity == "invalid", c("table", "var", "value", "reason")])
```

The report columns are: `table`, `var`, `row`, `haul.id`, `value`, `reason`,
`method` (`"rule"` or `"percentile"`), `severity` (`"invalid"` or `"extreme"`),
`p_lo` / `p_hi` (the percentiles used), `thr_lo` / `thr_hi` (the resulting
thresholds) and `group` (the grouping key).

Three further attributes hold the affected haul identifiers:

```{r}
c(all = length(attr(surv, "outlier_hauls")),
  invalid = length(attr(surv, "outlier_hauls_invalid")),
  extreme = length(attr(surv, "outlier_hauls_extreme")))
```

Which variables drive the flags:

```{r}
with(report, table(var, severity))
```

### Acting on the results

```{r}
## Remove only the rule-based failures; percentile flags are kept
cleaned <- check_outliers(surv, action = "remove", verbose = FALSE)
nrow(cleaned[["HH"]])
```

`remove_extremes = TRUE` additionally removes the percentile-flagged hauls, but
that is rarely appropriate: with `pct_probs = c(0.01, 0.99)`, roughly two per
cent of the data is flagged by construction, whether or not anything is wrong.
The flags are a place to start looking, not a verdict.

Narrower or stricter runs:

```{r}
## Only check the variables that feed the swept-area calculation
sa_check <- check_outliers(surv, vars = c("HaulDur", "Distance", "GroundSpeed",
                                          "DoorSpread", "WingSpread"),
                           pct = TRUE, pct_probs = c(0.005, 0.995),
                           verbose = FALSE)
nrow(attr(sa_check, "outlier_report"))
```

A threshold of your own is usually easier to apply to the report than to
express as a new rule. The report gives you the haul identifiers, so any
selection can be turned into a subset:

```{r}
## Hauls flagged for haul duration, by method
subset(report, var == "HaulDur")[c("value", "method", "thr_lo", "thr_hi")][1:5, ]

## Your own threshold: hauls shorter than 15 minutes
short_ids <- unique(report$haul.id[report$var == "HaulDur" &
                                   as.numeric(report$value) < 15])
length(short_ids)

## Drop them if you decide they are unusable
nrow(subset(surv, !haul.id %in% short_ids)[["HH"]])
```

## `check_lengths()`

Summarises the pooled length distribution of a single species and compares the
observed range with the empirical maximum length from the bundled
`species_info` table.

```{r, fig.height = 7, warning = FALSE}
plaice <- subset(surv, Valid_Aphia == 127137)
plaice <- check_lengths(plaice)
```

```{r}
attr(plaice, "length_check")$lPars
attr(plaice, "length_check")$nAbove
```

`maxEmp` is the published maximum length for the species (from FishBase); `perc`
is the requested upper percentile of the observed distribution
(`length_percentile`, default `99.9999`). `nAbove` gives the number and
percentage of observations above each. A non-trivial count above `maxEmp` points
at length-unit errors or misidentification. Use `cm_breaks` or `by` to control
the binning, and `plot = FALSE` to skip the figure.

## `check_weights()`

Fits a log-log length-weight relationship to the individual weights in `CA` and
compares it with the lookup parameters in `species_info`.

```{r, fig.height = 7, warning = FALSE}
plaice <- add_numbers_at_length(plaice)
plaice <- check_weights(plaice)
```

```{r}
rbind(fitted = attr(plaice, "weight_check")$parEst,
      lookup = attr(plaice, "weight_check")$parEmp)
```

A fitted exponent *b* far from 3, or a large discrepancy against the lookup
values, usually indicates unit errors or contaminated records. `max_length` and
`max_weight` exclude implausible observations before fitting:

```{r, fig.show = "hide", warning = FALSE}
plaice2 <- check_weights(plaice, max_length = 60, max_weight = 5000)
attr(plaice2, "weight_check")$parEst
```

## `add_swept_area()`

Swept area is where survey data are at their most incomplete, so the function
reports what it had to invent.

```{r}
surv_sa <- add_swept_area(surv, full_report = TRUE)
```

The summary is attached as an attribute and also gives a per-haul flag:

```{r}
head(attr(surv_sa, "swept_area_summary"))

## Share of hauls whose spread or distance was imputed
mean(surv_sa[["HH"]]$SweptArea_imputed)
```

Read the table by survey and gear rather than in aggregate. A gear with
`prop_NA` of 1 records neither of the required fields and cannot be given a
swept area at all; a gear with a high `prop_imputed` has swept areas built from
gear-level medians rather than haul-specific measurements. Both are legitimate,
but an abundance index computed over them carries very different uncertainty
than the haul count suggests.

The plausibility filters are adjustable: `min_speed` (default 1 knot),
`min_dist` (default 0) and `max_dist_dev` (default 0.2, the tolerated relative
difference between recorded distance and distance reconstructed from duration
and speed). `width = "DoorSpread"` switches the width source for surveys that
record door spread but not wing spread. `method = "fishglob"` uses the
survey-specific spread models of the FishGlob workflow instead.


# Stage 6: Inspecting the result

## Tabular summaries

`print()` and `summary()` double as QC reports. `summary()` picks up the
outlier attributes automatically:

```{r}
summary(surv)
```

Things worth reading off this output: whether any year is missing from the
series, whether the coordinate and depth ranges are plausible for the surveys
included, the ratio of zero-catch to positive hauls, and the outlier breakdown.

## Mapping the flagged hauls

The most efficient way to decide whether flagged records are a data problem or a
real feature is to look at where they are.

```{r, fig.height = 6, warning = FALSE}
bad <- subset(surv, haul.id %in% attr(surv, "outlier_hauls_invalid"))
plot_datras_overview(bad, col = "#A6081A", cex = 1.2)
```

Flags concentrated in one survey, one ship or one year point at a recording
convention rather than at individual bad hauls.

## Spatial and temporal coverage

```{r, fig.height = 6, warning = FALSE}
plot_datras_overview(surv, mode = "grid", metric = "count_hauls",
                     spatial_basis = "statrec")
```

```{r, fig.height = 7, warning = FALSE}
plot_datras_overview(surv, value_var = "Depth", by_survey = TRUE,
                     multi_panels = TRUE)
```

Mapping a covariate such as depth by survey exposes both spatial gaps and
values that are locally implausible even though they pass a global range check.

## Length and composition sanity

```{r, fig.height = 5, warning = FALSE}
plot_length_distribution(plaice, log_scale = TRUE)
```

A spectrum with isolated spikes at round numbers, or a long thin tail beyond the
empirical maximum length, indicates unit or transcription problems.

```{r, fig.height = 6, warning = FALSE}
comp <- add_numbers_at_length(surv)
comp <- add_total_numbers_by_haul(comp, length_cuts = c(0, 15, 30, 50, Inf))
plot_species_composition(comp, beside = FALSE, legend_ncol = 2)
```

Composition by length group exposes problems a total does not: a species that
appears only in one length group, or a length group dominated by a single
species across every survey, is usually a binning or identification artefact
rather than ecology.

## Spatial attribution

```{r}
surv_area <- add_ices_areas(surv)

## Adds Area_Full, Area_27 and Ecoregion to HH
head(sort(table(surv_area[["HH"]]$Area_27), decreasing = TRUE))

## Hauls that could not be attributed
sum(is.na(surv_area[["HH"]]$Area_27))
```

The function reports how many hauls it matched. Hauls that cannot be matched to
an ICES area usually have a coordinate problem that range checks alone will not
catch.


# Further checks worth considering

The following are not built into **DATRASextra**, because the right treatment
depends on the analysis. They are common sources of trouble.

**Survey design changes over time.** Gear, ship and sweep length changes create
step changes in catchability that look like population signal.

```{r}
with(surv[["HH"]], table(Year, Gear))
```

**Species validity codes.** `SpecVal` in `HL` marks how a species record should
be used, and `clean_datras()` does not filter on it. Code `1` is fully valid;
others indicate partial or unusable records.

```{r}
table(mini[["HL"]]$SpecVal, useNA = "ifany")
```

**Mixed reporting conventions.** Check `DataType` per survey and year, not just
overall (see Stage 3).

**Day and night composition.** Catchability of many species differs strongly
between day and night, and the day/night ratio can drift across years.

```{r}
with(surv[["HH"]], table(Year, DayNight))
```

**Seasonal drift.** A survey nominally in quarter 1 may drift by several weeks
over a time series; `timeOfYear` makes that visible.

```{r}
tapply(surv[["HH"]]$timeOfYear, surv[["HH"]]$Year, function(z)
  round(range(z, na.rm = TRUE), 3))
```

**Depth against bathymetry.** Recorded depth can be cross-checked against an
external bathymetry raster. Systematic offsets usually indicate a unit or datum
problem.

**Internal catch consistency.** `TotalNo` and `CatCatchWgt` in `HL` should be
consistent with the summed `HLNoAtLngt` for the same haul, species and category.
Large discrepancies indicate sub-sampling information that has not been applied
consistently.

**Age data plausibility.** Length-at-age should be monotone on average and its
spread should not collapse or explode at the oldest ages. Also check ALK
coverage: ages sampled in too few hauls give unstable keys.

**Duplicate biological records.** Exact duplicate rows within a haul and species
in `HL` or `CA` are worth counting, since they inflate abundance without
tripping any range check.

**Reference data coverage.** `add_species_info()` joins the bundled species
table; check which species come back without length-weight parameters or maximum
length, because those silently degrade any weight-based calculation.

```{r}
info <- add_species_info(surv, vars = c("maxL", "a", "b"), verbose = FALSE)

## TRUE means the species has no length-weight parameters in species_info
with(info[["HL"]], tapply(is.na(a), Species, all))
```

All five species in `mini` are covered. In a full survey extract, with several
hundred taxa, a handful typically are not.


# A reproducible recipe

Putting the stages together:

```{r, eval = FALSE}
## 1-2. Download once, then read from the archive
download_datras(path = "data/datras", surveys = "NS-IBTS", years = 2015:2023)
x <- read_datras("data/datras", surveys = "NS-IBTS", years = 2015:2023)

## Audit the read before anything is filtered
sum(is.na(x[["CA"]]$haul.id))
table(x[["HH"]]$DataType)
get_accuracy_cm(x)

## 3. Clean, with a full account of what is removed
x <- clean_datras(x, explain_codes = TRUE)

## 4. Check, without removing anything yet
x <- check_outliers(x, pct = TRUE)
report <- attr(x, "outlier_report")

## 5. Look at the flags before deciding
plot_datras_overview(subset(x, haul.id %in% attr(x, "outlier_hauls_invalid")))

## 6. Remove only what you have decided is wrong
x <- check_outliers(x, action = "remove", verbose = FALSE)

## 7. Derived variables, with their own diagnostics
x <- add_swept_area(x, full_report = TRUE)
x <- add_numbers_at_length(x)
x <- add_total_numbers_by_haul(x)

summary(x)
```


# Summary

| Stage             | Automatic steps                                                                                    | Under your control                                                                                                                           |
|-------------------|----------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------|
| Download          | code renaming, `-9` to `NA`, strict haul matching, exchange archiving                              | `surveys`, `years`, `download_hl`, `download_ca`, `include_flagged`, `overwrite`, `timeout`                                                  |
| Read              | corrupt-file check, small-file exclusion, duplicate haul removal, type-safe combining              | `strict`, `min_file_size`, `prune`, `drop_hl`, `drop_ca`, `surveys`, `years`, `ncores`                                                       |
| Derived variables | `haul.id`, `LngtCm`, `Species`, `Count`, `lon`/`lat`, `abstime`, `Roundfish`, zero-catch treatment | length binning via `cm_breaks` / `by`                                                                                                        |
| Cleaning          | `HaulVal` and `StdSpecRecCode` filters                                                             | `verbose`, `explain_codes`, `impute_missing_depth`, `correct_species`, `aphias`, `years`, `quarters`, `gears`, `do_fishglob`                 |
| Checks            | none by default                                                                                    | `check_outliers()` (`strict`, `pct`, `pct_*`, `vars`, `action`, `remove_extremes`), `check_lengths()`, `check_weights()`, `add_swept_area()` |

Nothing in the automatic column is hidden: the download and read stages print
what they do, `clean_datras()` reports the data before filtering, and every
check writes its findings to an attribute you can inspect. Where a step cannot
be switched off by argument, it is a plain `subset()` call that can be replaced
with a rule of your own.

```{r, echo=FALSE}
par(oldpar)
```


# References

- ICES DATRAS database: <https://datras.ices.dk>
- ICES vocabularies for `HaulVal`, `StdSpecRecCode` and `BySpecRecCode`:
  <https://vocab.ices.dk>
- `vignette("datrasextra-tutorial")` for the basic workflow
- `vignette("articles/fishglob")` for the FishGlob pipeline
