---
title: "Introduction to DATRASextra"
description: >
  Learn how to get started with the basics of DATRASextra.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to DATRASextra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```


The goal of the DATRASextra is to provide useful additional functions to work
with [DTU Aqua's *DATRAS* R package](https://github.com/DTUAqua/DATRAS) and
[ICES' DATRAS
database](https://www.ices.dk/data/data-portals/pages/datras.aspx). To get
started, make sure to install the most recent package version from GitHub.


## Installation

The current version of *DATRASextra* (v0.0.1) requires R $>= 4.0.0$ and can be
downloaded from github as follows.

::: .momo-release
```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("tokami/DATRASextra")
```
:::

<!-- The development version can be installed with: -->

<!-- ::: .momo-devel -->
<!-- ```{r, eval = FALSE} -->
<!-- remotes::install_github("tokami/DATRASextra", ref = "dev") -->
<!-- ``` -->
<!-- ::: -->


## Basic usage

After successful installation, the package can be loaded into the R environment.

```{r, fig.alt = ""}
library(DATRASextra)
```

As a first step, we need to download the DATRAS database. The database contains
information of more than 27 surveys. You can get some information in which years and
quarters the surveys operated with:

```{r, fig.alt = ""}
knitr::kable(listSurveys())
```

The `downloadDATRAS` function allows us to download information for any of these
surveys. If the arguments `surveys` and `years` are not specified (`NULL`), all
surveys and years are downloaded, wich can take a while. For demonstration
purposes, we download year 2023 of the NS-IBTS survey into a temporary
directory:

```{r, fig.alt = ""}
survey <- "SNS"
tmp <- tempdir()
downloadDATRAS(surveys = survey, years = 2023, dir = tmp)
```

If no directory (argument `dir`) is specified, the data is downloaded into the
current working directory (`getwd()`). The function compresses downloaded data
into "zip" archives. Note, that if a year is specified in which a specific
survey did not operate, an empty file is created in the specified folder.
However, the next function just reads in non-empty files.

Next, we can read the downloaded survey information into R:

```{r, fig.alt = ""}
surv0 <- readDATRAS(file.path(tmp, survey))
```

The path can point to a specific file that you want to read in or to a whole
directory with any number of files.

Now, the `surv0` object contains the survey information in the usual `DATRASraw`
format of the *DATRAS* package. (TODO: Link to more information about this data
type or provide more information here.) It make sense to clean the data and for
example only use valid hauls. This can be done manually with the `subset()`
function or by using some recommended cleaning steps with the `clean()`
function. This function also allows us to make a relevant subset of the database
based on our needs.

```{r, fig.alt = ""}
surv <- clean(surv0)
```

By default, this function also imputes missing depth information.

We can plot the resulting survey information with:

```{r, fig.alt = ""}
plot(surv)
```


## A more detailed tutorial

The [tutorial vignette](tutorial.html) provides a step-by-step guide to applying
*DATRASextra* to download, read-in, and process ICES DATRAS database using [DTU
Aqua's *DATRAS* R package](https://github.com/DTUAqua/DATRAS).
