---
title: "A Step-by-step guide to working with the ICES DATRAS database using DATRASextra"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{A Step-by-step guide to working with the ICES DATRAS database using DATRASextra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7
)
```

The goal of the **DATRASextra** is to make it easier to work with the [ICES
DATRAS database](https://www.ices.dk/data/data-portals/pages/datras.aspx) in R.
The package provides user-friendly, well-document functions, detailed vignettes,
and functionality that extends the powerful core tools of the [**DATRAS** R
package](https://github.com/DTUAqua/DATRAS).

This vignette introduces a typical workflow for working with ICES DATRAS survey
data, from installing **DATRASextra** and downloading survey information to
reading, cleaning, summarising, and plotting the data.


## Outline

1. Install **DATRASextra**
2. Download ICES DATRAS survey data
3. Read downloaded data into R
4. Clean and inspect the data
5. Calculate total numbers and weight by haul
6. Summary
7. References


## Install DATRASextra

To get started, install the latest development version of **DATRASextra** from
GitHub:

```{r, eval = FALSE, echo = TRUE, fig.alt = ""}
## Install the package
remotes::install_github("tokami/DATRASextra")
```

Then load the package into your R session:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Load the package
library(DATRASextra)
```



# Download DATRAS survey data

The DATRAS database is a large compilation of scientific bottom-trawl survey
data from across the Northeast Atlantic. You can get an overview of the
available surveys with:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## List available DATRAS surveys
list_surveys()
```

Or visually with:

```{r, eval = TRUE, echo = TRUE, fig.alt = "Map with available surveys in DATRAS"}
## Plot available DATRAS surveys
plot_datras_overview(by_survey = TRUE, multi_panels = TRUE)
```


The `download_datras()` function can be used to download data for one or more
surveys. If both `surveys` and `years` are left as `NULL`, all available surveys
and years are downloaded. Because the database is large, a full download can
take some time.

In many cases, however, only a subset of the database is needed. The `surveys`
and `years` arguments allow you to restrict the download to selected surveys and
years. For example, the 2023 `SNS` survey can be downloaded into a temporary
directory with:

```{r, eval = FALSE, echo = TRUE, fig.alt = ""}
## Select survey
survey <- "SNS"

## Create temporary directory
tmp <- tempdir()

## Download SNS data for 2023
download_datras(surveys = survey, years = 2023, dir = tmp)
```

Note that `download_datras()` does not load the data directly into R. Instead,
it downloads, compresses, and stores the files on disk. If no directory is
specified via `dir`, the files are saved to the current working directory, which
you can inspect with `getwd()`.

Downloaded files are stored as zipped DATRAS exchange files. If a year is
requested for which a survey was not conducted, an empty file may be created.
The reading functions described below automatically ignore empty files.

DATRAS typically contains three main data tables:

- `HH`: haul- or station-level metadata, such as position, time, gear, and
  environmental information
- `HL`: species-level catch-at-length and subsampling data
- `CA`: individual-level biological data, such as length, weight, sex, maturity,
  and age

The `CA` table can be large and may not be needed for all analyses. In such
cases, setting `download_ca = FALSE` can reduce download time and memory use.
Similarly, when the focus is on individual biological data rather than
catch-at-length information, `HL` can be omitted by setting `download_hl =
FALSE`. See the `vignette("articles/exploring-length-at-age")` for an example when working only with CA
might be useful.


# Read DATRAS data into R

The downloaded survey files can then be read into R with `read_datras()`:

```{r, fig.alt = "", eval = FALSE}
surv0 <- read_datras(file.path(tmp, survey))
```

The path can point either to a single file or to an entire directory containing
multiple survey-year files.

To avoid relying on an internet connection for the remainder of this vignette,
we use the example data set `dab` included in **DATRASextra**:

```{r, fig.alt = ""}
surv0 <- dab
```

# Clean and inspect DATRAS data

The object returned by `read_datras()` has class `datras_raw` and is
intentionally kept close to the original DATRAS exchange format. In many
analyses it is useful to clean the data first, for example by restricting the
data set to valid hauls only.

This can be done manually with `subset()`, but **DATRASextra** also provides the
convenience function `clean_datras()` for applying common cleaning steps and
creating relevant subsets:

```{r, fig.alt = ""}
surv <- clean_datras(surv0)
```

Another useful feature of `clean_datras()` is that it can impute missing depth
information where appropriate.

After cleaning, it is often sensible to check for invalid or unusual values.
This can be done with `check_outliers()`:

```{r, fig.alt = ""}
surv <- check_outliers(surv, pct = TRUE)
```

The function returns a summary of hauls with invalid values and, when `pct =
TRUE`, also flags unusually extreme values in selected variables.

More detailed information is available in the attributes added by the function:

```{r, fig.alt = ""}
head(attr(surv, "outlier_report"))
```

A shorter list of affected hauls can be obtained with:

```{r, fig.alt = ""}
head(attr(surv, "outlier_hauls"))
```

For the example data set used here, no clearly invalid values were detected, so
we proceed without further filtering.

A quick overview of the cleaned survey data can be obtained with:

```{r, fig.alt = ""}
plot(surv)
```


# Calculate total numbers and weight by haul

A commonly used quantity in survey analysis is the total abundance or biomass
per haul. These values are often used to derive survey indices, community
indicators, or biodiversity metrics.

To calculate total numbers and weight by haul, the information in the `HL` table
first needs to be raised to the haul level. A convenient first step is to add
numbers-at-length:

```{r, fig.alt = ""}
surv <- add_numbers_at_length(surv)
```

This adds a matrix `N` to the `HH` table, containing the estimated numbers in
each length class for each haul. The length classes can be modified with the
`cm_breaks` or `by` arguments.

The total number by haul can then be obtained by summing over length classes:

```{r, fig.alt = ""}
surv <- add_total_numbers_by_haul(surv)
```

By default, this adds a single column `HaulN` to the `HH` table. More generally,
custom bins can be specified via `length_cuts`. In that case, the result may
contain separate totals for broader user-defined length groups.

If species-specific length and weight information are available in `CA`, or
suitable empirical length-weight relationships are available in the
`species_info` table, the numbers can also be converted to weight:

```{r, fig.alt = ""}
surv <- add_total_weight_by_haul(surv)
```

This adds `HaulWgt` to the `HH` table, containing the total estimated weight by
haul. As for abundance, custom length-group summaries can be generated when
`length_cuts` is used.

# Summary

This vignette introduced a basic workflow for working with ICES DATRAS survey
data in **DATRASextra**. It showed how to

- install and load the package,
- inspect available DATRAS surveys,
- download selected survey data,
- read DATRAS exchange files into R,
- clean and check the data,
- visualise the resulting survey information, and
- derive haul-level abundance and biomass summaries.

These functions provide a practical starting point for exploratory analyses,
survey summaries, and the preparation of DATRAS data for downstream applications
such as indicator development, biodiversity analyses, and stock assessment.

# References
