---
title: "Step-by-step guide to work with the ICES' DATRAS database"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Step-by-step guide to work with the ICES' DATRAS database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

The goal of the **DATRASextra** is to make it easy to work with [ICES' DATRAS
database](https://www.ices.dk/data/data-portals/pages/datras.aspx) by providing
user-friendly, well-document functions, detailed vignettes, and building on the
powerful [*DATRAS* R package](https://github.com/DTUAqua/DATRAS).

This vignette sets you up for working with ICES DATRAS data base by taking you
through the whole process from installing *DATRASextra* and downloading DATRAS,
to summarizing and plotting observations from the scientific surveys.


**Outline:**

1. Install DATRASextra
2. Download ICES DATRAS
3. Read-in DATRAS
4. Clean DATRAS
5. Get Weight by haul
6. Summary
7. References



# *Install DATRASextra*

To get started, make sure to install the most recent package version from
GitHub:

```{r, eval = FALSE, echo = TRUE, fig.alt = ""}
## Install the package
remotes::install_github("tokami/DATRASextra")
```

The package can be loaded into the R environment:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Load the package into R
library(DATRASextra)
```


# *Download DATRAS*

The DATRAS data base is a large compilation of the majority of scientific
bottom-trawl surveys in the Northeast Atlantic. You can get an overview of
included surveys with:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## List scientific surveys in DATRAS
list_surveys()
```

The `download_datras` function allows you to download information for any of
these surveys. If the arguments `surveys` and `years` are not specified (i.e.,
`NULL`), all surveys and years are downloaded, wich can take a while (~
30-60min).

If not the whole database is required, the `surveys` and `years` arguments allow
to specify specific surveys and years. For example, the `Sole Net Survey (SNS)`
in 2023 could be downloaded to a temporary directory with:

```{r, eval = FALSE, echo = TRUE, fig.alt = ""}
## select survey
survey <- "SNS"
## create temporary directory
tmp <- tempdir()
## download SNS for 2023
download_datras(surveys = survey, years = 2023, dir = tmp)
```

If no directory (argument `dir`) is specified, the data is downloaded into the
current working directory (use `getwd()` to see your current working directory).
The function compresses downloaded data into "zip" archives. Note, that if a
year is specified in which a specific survey did not operate, an empty file is
created in the specified folder. However, the next function just reads in
non-empty files.

DATRAS includes three main data sets, `HH` with information on haul level, `HL`
with information on species level and the important information on numbers by
length, and `CA` with information on the individual level, such as individual
weight, age readings, or maturity scores. The third data set with the biological
information (`CA`) can be quite large and might not be required for some
projects. In that case, setting the argument `download.ca = FALSE` allows you to
to omit this data set from the download and save some time and memory.


# *Read-in DATRAS*

Next, we can read the downloaded survey information into R:

```{r, fig.alt = "", eval = FALSE}
surv0 <- read_datras(file.path(tmp, survey))
```

The path can point to a specific file that you want to read in or to a whole
directory with any number of files.


# *Clean DATRAS*


Now, the `surv0` object contains the survey information in the usual `datras_raw`
format of the *DATRAS* package. (TODO: Link to more information about this data
type or provide more information here.) It make sense to clean_datras the data and for
example only use valid hauls. This can be done manually with the `subset()`
function or by using some recommended cleaning steps with the `clean_datras()`
function. This function also allows us to make a relevant subset of the database
based on our needs.

```{r, fig.alt = ""}
data("dab")
surv0 <- dab
```


```{r, fig.alt = ""}
surv <- clean_datras(surv0)
```

By default, this function also imputes missing depth information.


We can plot the resulting survey information with:

```{r, fig.alt = ""}
plot(surv)
```



# *Get Weight by haul*




# *Summary*

This vignette introduced the main features and workflow of the *DATRASextra*
package for ...



# *References*
