---
title: "Numbers and weight by haul and length classes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Numbers and weight by haul and length classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

The goal of this vignette is to demonstrate how to derive numbers and weight by
haul and length classes based on the DATRAS database. This requires following
steps:

**Outline:**

1. Create length spectrum
2. Create weight spectrum
3. Aggregated length classes
4. Summary
5. References


The vignette uses the the *DATRASextra* package:

```{r}
library(DATRASextra)
```

and assumes that a cleaned *DATRAS* database is available (see tutorial
vignette) for information on how to get a cleaned *DATRAS* file. Here, we use
the dab data set included in the package.

## *Create length spectrum*

```{r}
lpars <- check_lengths(dab)
```



```{r}
dab <- add_numbers_at_length(dab)
```


Alternatively, custom length breaks with e.g. maximum length threshold:


```{r}

## TODO: ugly code, can this be improved with some helper functions?

l.cut.off <- lpars$lPars$median * 5  ## find good cut-off
## or l.cut.off <- lpars$lPars$perc
cm.breaks <- attr(lpars$N,"cm.breaks")
nbyl <- colSums(lpars$N)[which(cm.breaks <= l.cut.off)]
threshold <- cm.breaks[max(which(nbyl != 0))]
```





```{r}

## TODO: ugly code, can this be improved with some helper functions?

## add length data with custom breaks
by <- get_accuracy_cm(dab)
cm_breaks <- seq(lpars$lPars$min, threshold + by, by = by)
LngtCode2cm <- c(. = 0.1, `0` = 0.5, `1` = 1, `2` = 2, `5` = 5)
while (length(cm_breaks) > 100 && which(by == LngtCode2cm) != length(LngtCode2cm)) {
    by <- LngtCode2cm[which(by == LngtCode2cm) + 1]
    cm_breaks <- seq(lpars$lPars$min, threshold + by, by = by)
}

dab <- add_numbers_at_length(dab, cm_breaks = cm_breaks, by = by)
```




## *Convert numbers at length into weight at length*


```{r}
wpars <- check_weights(dab)
```


```{r}
dab <- add_weight_at_length(dab, max_length = threshold)
```


Alteratively, ...

```{r}
# dab <- add_weight_at_length(dab, empirical = TRUE)
```




## *Aggregated length classes*

If for example only 2 or 3 length classes: ....

```{r}
## TODO: function to get length and weight ranges from datras_raw object

length_cuts <- c(0, 10, 20, 30, 100)

dab <- add_total_numbers_by_haul(dab, length_cuts = length_cuts)
dab <- add_total_weight_by_haul(dab, length_cuts = length_cuts)
```


HaulN and HaulWgt are now provided as matrices with numbers and weight over the
specified length classes, respectively.


Plot:


```{r}
ncols <- ncol(dab[["HH"]][["HaulWgt"]])
par(mfrow = n2mfrow(ncols, asp = 2),
    mar = c(3,3,2,2), oma = c(2,2,0,0))
for (i in 1:ncols) {
    plot(dab[["HH"]]$lon, dab[["HH"]]$lat,
         ty = "n",
         xlab = "", ylab = "",
         main = colnames(dab[["HH"]]$HaulN)[i])
    ind <- which(dab[["HH"]]$HaulN[,i] > 0)
    points(dab[["HH"]]$lon[ind], dab[["HH"]]$lat[ind],
           cex = dab[["HH"]]$HaulN[,i][ind] /
               max(dab[["HH"]]$HaulN[,i][ind]) * 3,
           col = i)
}
mtext("Longitude", 1, outer = TRUE)
mtext("Latitude", 2, outer = TRUE)
```




## *Summary*

This vignette introduced the main features and workflow of the *DATRASextra*
package for ...



## *References*
