---
title: "Deriving numbers and weight by haul and custom length classes"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Deriving numbers and weight by haul and custom length classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7
)
```

This vignette demonstrates how to derive numbers and weight by haul from ICES
DATRAS survey data using **DATRASextra**. It shows how to

- create a haul-by-length matrix of numbers,
- convert numbers at length to weight at length,
- aggregate these quantities over custom length classes, and
- derive haul-level summaries for all fish combined or for user-defined length
  groups such as juveniles and adults.

These summaries are useful for survey standardisation, indicator development,
biodiversity analyses, and stock-assessment workflows.

## Outline

1. Load **DATRASextra**
2. Create numbers at length
3. Convert numbers at length to weight at length
4. Aggregate numbers and weight over custom length classes
5. Summary
6. References

# Load DATRASextra

Load the package with:

```{r}
library(DATRASextra)
```

This vignette assumes that a cleaned `DATRAS` data set is already available. For
a full workflow covering download, reading, and cleaning of survey data, see
`vignette("datrasextra-tutorial")`.

Here, we use the example data set `dab` included in the package and apply a
standard cleaning step first:

```{r}
dab <- clean_datras(dab)
```

# Create numbers at length

Catch-at-length information is stored in the `HL` component of a `DATRAS`
object. Before aggregating these data, it is useful to inspect the available
length measurements:

```{r}
dab <- check_lengths(dab)
```

This function provides summary statistics and diagnostic plots of the recorded
lengths. It can help identify potential data issues such as implausible values
or unusual length distributions. If the data look reasonable, the next step is
to aggregate the `HL` observations to numbers at length by haul and add the
result to the `HH` component:

```{r}
dab <- add_numbers_at_length(dab)
```

This creates a matrix `HH$N`, with one row per haul and one column per length
class:

```{r}
dab[["HH"]][["N"]][1:5, 1:5]
```

By default, `add_numbers_at_length()` defines length classes from the minimum
and maximum observed lengths, using the coarsest length recording resolution
present in the data (see `?get_accuracy_cm`). In most cases, this is a sensible
default.

If a different resolution or range is needed, custom breaks can be supplied via
`cm_breaks` and `by`. In general, it is recommended to keep the numbers-at-
length matrix as fine as possible, because these length classes are also used to
derive weights at length in the next step. Coarser groupings can then be
obtained later with `add_total_numbers_by_haul()` or
`add_total_weight_by_haul()`.

For example, the following code creates numbers at length using 0.5 cm bins:

```{r}
## Define custom resolution of length bins
by <- 0.5

## Define custom length bins
cm_breaks <- seq(attr(dab, "length_check")$lPars$min, attr(dab, "length_check")$lPars$maxEmp, by = by)

## Recalculate numbers at length using custom bins
dab <- add_numbers_at_length(dab, cm_breaks = cm_breaks, by = by)
```

The resulting matrix now has a finer length resolution:

```{r}
dab[["HH"]][["N"]][1:5, 1:5]
```

# Convert numbers at length to weight at length

Numbers at length are sufficient for some applications, but many analyses also
require biomass or weight by length class or by haul. For example, biomass-based
survey indices are often used in stock assessment or ecosystem analyses.

To derive weights, **DATRASextra** uses the individual length and weight records
stored in the `CA` component to estimate a length-weight relationship. Before
doing so, it is useful to inspect the available weight data:

```{r}
dab <- check_weights(dab)
```

This function provides summary statistics and diagnostic plots for the weight
data and the fitted length-weight relationship. If the information looks
reasonable, the numbers-at-length matrix in `HH$N` can be converted to weight at
length with:

```{r}
dab <- add_weight_at_length(dab)
```

This adds a matrix `HH$Wgt`, containing the estimated weight in each length
class for each haul.

The function uses the midpoints of the existing length bins in `HH$N` (stored in
`attr(dab, "cm.breaks")`) and converts these lengths to weights using the fitted
length-weight relationship from the `CA` data.

One important caveat concerns the largest length class. If the final bin is a
plus group with a very large upper limit, such as `Inf`, or extends far beyond
the observed size range, the midpoint of that bin may imply an unrealistically
large weight. In such cases, it may be preferable to use the `plus_group`
argument.

If the `CA` component does not contain enough information to estimate a reliable
length-weight relationship, lookup parameters can instead be taken from the
`species_info` table by setting `lw_source = "lookup"`. Before doing so, it is good
practice to inspect the stored parameters:

```{r}
species_info[
  which(species_info$ScientificName_WoRMS == "Limanda limanda"),
  c("a", "b")
]
```

If these values appear appropriate, the lookup relationship in the species_info
table can be used with:

```{r eval = FALSE}
dab <- add_weight_at_length(dab, lw_source = "lookup")
```

# Aggregate numbers and weight over custom length classes

By default, `add_total_numbers_by_haul()` and `add_total_weight_by_haul()` sum
over all available length classes and return one total value per haul. In many
applications, however, it is useful to retain separate haul-level summaries for
custom length groups, for example juveniles and adults.

For dab, one biologically meaningful split is the approximate length at 50%
maturity, available in `species_info`:

```{r}
(lm <- species_info[
  which(species_info$ScientificName_WoRMS == "Limanda limanda"),
  "Lm"
])
```

We can use this value to define two custom groups, below and above `Lm`:

```{r}
length_cuts <- c(0, lm, Inf)

dab <- add_total_numbers_by_haul(dab, length_cuts = length_cuts)
dab <- add_total_weight_by_haul(dab, length_cuts = length_cuts)
```

This returns haul-level summaries for juveniles and adults separately:

```{r}
head(dab[["HH"]][["HaulN"]])
```

```{r}
head(dab[["HH"]][["HaulWgt"]])
```

By default, the column names reflect the chosen cut points, but they can be
renamed as needed:

```{r}
colnames(dab[["HH"]][["HaulN"]]) <- c("juveniles", "adults")
colnames(dab[["HH"]][["HaulWgt"]]) <- c("juveniles", "adults")
```

This makes the result easier to interpret:

```{r}
head(dab[["HH"]][["HaulN"]])
```

Of course, any number of custom length groups can be defined. For example:

```{r}
length_cuts <- c(0, 10, 20, 30, 100)

dab <- add_total_numbers_by_haul(dab, length_cuts = length_cuts)
dab <- add_total_weight_by_haul(dab, length_cuts = length_cuts)
```

This gives haul-level numbers and weights for four user-defined length groups:

```{r}
head(dab[["HH"]][["HaulN"]])
head(dab[["HH"]][["HaulWgt"]])
```

These summaries can then be used to explore whether the spatial distribution
differs among size groups. For example, the following plot shows where hauls
with positive catches occurred for each length class, with point size scaled by
haul-level abundance:

```{r}
ncols <- ncol(dab[["HH"]][["HaulWgt"]])

par(mfrow = n2mfrow(ncols, asp = 2),
    mar = c(3, 3, 2, 2),
    oma = c(2, 2, 0, 0))

for (i in seq_len(ncols)) {
  plot(dab[["HH"]]$lon, dab[["HH"]]$lat,
       type = "n",
       xlab = "", ylab = "",
       main = colnames(dab[["HH"]]$HaulN)[i])

  ind <- which(dab[["HH"]]$HaulN[, i] > 0)

  points(dab[["HH"]]$lon[ind], dab[["HH"]]$lat[ind],
         cex = dab[["HH"]]$HaulN[, i][ind] /
           max(dab[["HH"]]$HaulN[, i][ind]) * 3,
         col = i)
}

mtext("Longitude", 1, outer = TRUE)
mtext("Latitude", 2, outer = TRUE)
```

# Summary

This vignette showed how to derive haul-level abundance and biomass summaries
from DATRAS data using **DATRASextra**. Starting from a cleaned survey data set,
it demonstrated how to inspect the available length and weight information,
construct a numbers-at-length matrix, convert numbers at length to weight at
length, and aggregate both quantities over custom length classes.

These tools make it straightforward to generate biologically meaningful haul-
level summaries, either for the full size range or for selected groups such as
juveniles and adults. The resulting variables can be used directly in further
analyses, for example when building survey indices, comparing size-structured
spatial patterns, or deriving indicators based on abundance or biomass.

# References
