---
title: "The datras_raw object"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{The datras_raw object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7
)
```

This article describes the `datras_raw` object, the data structure used
throughout **DATRASextra**. We use the example data set `dab` (dab,
*Limanda limanda*, from NS-IBTS in 2020-2023).

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Load the package
library(DATRASextra)

## Load the example data
data(dab)
```


# Structure

A `datras_raw` object is a list of three data frames, with classes
`datras_raw` and `DATRASraw`:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Class and elements
class(dab)
names(dab)
```

The `DATRASraw` class is kept for backward compatibility with the [**DATRAS** R
package](https://github.com/DTUAqua/DATRAS), so its functions keep working on a
`datras_raw` object.

The three tables follow the DATRAS exchange format:

- `HH`: haul data. One row per haul (position, time, depth, gear,
  environment).
- `HL`: length data. Catch-at-length per species and length class.
- `CA`: age-length data. Individual length, weight, sex, maturity, age.

They are linked by `haul.id`, which is present in all three tables.

Printing the object gives a survey overview:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Survey overview
dab
```


# Indexing

Use `[[ ]]` to reach a table:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## The haul table
hh <- dab[["HH"]]
dim(hh)
```

Within a table, columns are accessed and assigned with `$` as usual:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## A single column and a few columns
head(dab[["HH"]]$Depth)
head(dab[["HH"]][, c("Year", "Quarter", "Gear", "Depth")])

## Add or modify a column
dab[["HH"]]$DepthLog <- log(dab[["HH"]]$Depth)
head(dab[["HH"]]$DepthLog)
```

`subset()` filters the whole object and keeps the three tables consistent
(dropping a haul from `HH` also drops its `HL` and `CA` records):

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Keep valid hauls from quarter 1
q1 <- subset(dab, Quarter == 1 & HaulVal == "V")
q1
```


# Coded variables (ICES vocabulary)

Many variables are stored as codes (`Gear`, `Country`, `Ship`, `StatRec`),
and species as a WoRMS AphiaID in `Valid_Aphia`:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Gear codes and species AphiaID
levels(factor(as.character(dab[["HH"]][["Gear"]])))
unique(dab[["HL"]][["Valid_Aphia"]])
```

Definitions for these codes are in the [ICES
vocabulary](https://vocab.ices.dk/) and the [DATRAS data
resources](https://datras.ices.dk/Data_products/DataResources.aspx). The
`species_info` table also maps AphiaIDs to scientific names and ecological
groups.


# Numbers-at-length: the `N` matrix

`add_numbers_at_length()` raises the `HL` counts to the haul level and stores
the result as a matrix `N` in `HH`, one row per haul and one column per length
class:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Add numbers-at-length
dab <- add_numbers_at_length(dab)

## One row per haul, one column per length class
## (showing the five largest catches and a few length classes)
big <- head(order(rowSums(dab[["HH"]][["N"]]), decreasing = TRUE), 5)
dab[["HH"]][["N"]][big, 13:17]
```

Length classes can be changed with `cm_breaks` or `by`; see
`vignette("articles/custom-length-classes")`.


# Weight-at-length: the `Wgt` matrix

`add_weight_at_length()` converts numbers to weight using a length-weight
relationship and adds a matrix `Wgt` to `HH` with the same shape as `N`:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Add weight-at-length
dab <- add_weight_at_length(dab)

## Same hauls and length classes as above, now in weight
dab[["HH"]][["Wgt"]][big, 13:17]
```

Summing these matrices over their columns gives total numbers and weight per
haul, via `add_total_numbers_by_haul()` and `add_total_weight_by_haul()` (see
the [step-by-step guide](../articles/datrasextra-tutorial.html)).
