---
title: "Extracting Length-at-Age (LAA) from DATRAS"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Extracting Length-at-Age (LAA) from DATRAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7
)
```

This vignette demonstrates how to extract and explore length-at-age (LAA) data
from a cleaned `DATRAS` object and how to estimate stochastic age-length keys
(ALKs) using **DATRASextra**.

Length-at-age information is useful in its own right, for example for exploring
temporal, spatial, or seasonal patterns in growth and population structure.
However, it is also central to many stock-assessment workflows because the
individual biological observations in `CA` are typically *not*
catch-representative on their own. In contrast, the `HL` table contains the
representative catch-at-length information. ALKs are therefore used to convert
representative catch-at-length data into representative catch-at-age estimates.

In other words:

- `HL` tells us how many fish were caught at each length,
- `CA` tells us how age relates to length for the aged fish,
- ALKs combine both sources of information to infer catch at age.

## Outline

1. Load **DATRASextra**
2. Select years and clean the data
3. Explore length-at-age data
4. Estimate age-length keys (ALKs)
5. Predict catch at age from the fitted ALK
6. Summary

# Load DATRASextra

Load the package with:

```{r}
library(DATRASextra)
```

A typical `DATRAS` object is a list with three core components. We use the
example data set `dab` included in **DATRASextra** to demonstrate the workflow:

```{r}
str(dab, max.level = 1)
```

The three main tables are:

- `CA`: individual-level biological data, such as length, weight, sex,
  maturity, and age
- `HH`: haul- or station-level metadata, such as position, time, gear, and
  environmental information
- `HL`: catch-at-length and subsampling information

When a species appears in `CA`, a corresponding record should also exist in
`HL`, even if some values are missing or represented by special codes such as
`-9`.

# Select years and clean the data

In many applications, it is useful to restrict the data to a selected time
period before exploring LAA patterns or fitting an ALK. Here, we focus on the
years 2020 to 2024:

```{r}
years <- 2020:2024
```

We then apply a standard cleaning step with `clean_datras()`:

```{r}
dab <- clean_datras(dab, years = years)
```

This function applies a number of common filters and harmonisation steps. The
exact defaults may depend on the package version and settings, but the general
aim is to create a cleaner, more analysis-ready data set.

Equivalent manual filtering would look like this:

```{r eval=FALSE}
dab <- subset(
  dab,
  Year %in% years &
    HaulVal == "V" &
    DayNight == "D" &
    StdSpecRecCode == 1
)
```

Using `clean_datras()` is generally more convenient and helps keep workflows
consistent and reproducible.

# Explore length-at-age data

We start by extracting the `CA` table:

```{r}
ca <- dab[["CA"]]

## Structure of CA data
str(ca, max.level = 1)
```

The `CA` component contains the individual-level biological observations,
including age where available. This is the main source of information for
estimating age-length relationships.

## Sample size by year

A first useful check is to inspect how much data are available by year:

```{r}
## Number of individual rows per year
xtabs(~ Year, ca)

## Number of unique hauls per year
xtabs(~ Year, data = unique(ca[c("Year", "haul.id")]))
```

This gives a quick impression of the temporal distribution of the biological
sampling effort. Strong imbalances among years can matter when interpreting
temporal patterns or when fitting more flexible ALK models.

## Add spatial information from HH

Some useful variables, such as haul position, are stored in `HH` rather than
`CA`. To explore the spatial distribution of aged fish, we merge this
information onto the `CA` table using `haul.id`:

```{r}
ca <- merge(
  ca,
  dab[["HH"]][c("haul.id", "lon", "lat")],
  by = "haul.id",
  all.x = TRUE,
  sort = FALSE,
  suffixes = c("", ".HH")
)
```

We can now inspect where age observations were collected. The following plot
shows the haul locations of fish assigned to different ages:

```{r fig.width=6, fig.height=5}
ages <- sort(unique(ca$Age))
ages <- ages[!is.na(ages)]

xlim <- range(ca$lon, na.rm = TRUE)
ylim <- range(ca$lat, na.rm = TRUE)

par(mfrow = n2mfrow(length(ages)),
    mar = c(3, 3, 1, 1))

for (i in seq_along(ages)) {
  plot(ca$lon[ca$Age == ages[i]],
       ca$lat[ca$Age == ages[i]],
       col = i,
       xlim = xlim,
       ylim = ylim,
       xlab = "", ylab = "")

  plot(sf::st_geometry(DATRASextra:::get_land()), add = TRUE,
       col = grDevices::adjustcolor(grey(0.9), 0.4),
       border = grDevices::adjustcolor(grey(0.6), 0.4))

  legend("topleft", legend = paste0("Age: ", ages[i]),
         pch = NA, bg = "white")
  box(lwd = 1.5)
}
```

This type of plot can be helpful for spotting strong spatial clustering of age
samples. Such patterns may suggest that age data are concentrated in specific
areas, which can motivate the use of spatial terms in the ALK model.

## Age availability

The next step is to check how many observations actually have age information:

```{r}
table(ca$Age, useNA = "ifany")
```

```{r}
sum(is.na(ca$Age))
```

Around `r round(mean(is.na(ca$Age)) * 100, 1)`% of `CA` entries have missing
age.

This is expected in many DATRAS data sets: not every measured fish is aged. That
is precisely why ALKs are needed.

For ALK estimation, the variable `NoAtALK` is used as the frequency or count
associated with each age-length observation. Depending on the survey and data
structure, one row in `CA` may therefore represent one or multiple fish in the
age-length sample.

## Explore the age-length relationship directly

Before fitting a model, it is often useful to visualise the observed age-length
relationship:

```{r}
plot(ca$LngtCm, ca$Age,
     pch = 16, cex = 0.6,
     xlab = "Length [cm]",
     ylab = "Age")
box(lwd = 1.5)
```

This quick plot can reveal whether age tends to increase smoothly with length,
whether there is substantial overlap among ages, and whether outliers or unusual
records may require further checking.

A year-specific view can also be informative:

```{r}
par(mfrow = n2mfrow(length(unique(ca$Year))),
    mar = c(3, 3, 2, 1))

for (yr in sort(unique(ca$Year))) {
  if (any(!is.na(ca$Age[ca$Year == yr]))) {
    plot(ca$LngtCm[ca$Year == yr],
         ca$Age[ca$Year == yr],
         pch = 16, cex = 0.6,
         xlab = "", ylab = "",
         main = yr)
    box(lwd = 1.5)
  }
}
mtext("Length [cm]", side = 1, outer = TRUE)
mtext("Age", side = 2, outer = TRUE)
```

If the relationship changes substantially among years, a model with year effects
may be preferable. As the dab example shows, for some years, age information
might not be available (here: 2020 and 2022).

# Estimate age-length keys (ALKs)

To predict catch at age, we first need representative catch-at-length
information from `HL`. We therefore add the numbers-at-length matrix to the
object:

```{r}
dab <- add_numbers_at_length(dab)
```

For stock-assessment applications, ALKs are often estimated separately by
quarter. As a simple example, we restrict the data to quarter 1:

```{r}
dab_Q1 <- subset(dab, Quarter == 1)
```

We can now fit a basic stochastic ALK:

```{r}
ALK <- fitALK(
  dab_Q1,
  minAge = 3,
  maxAge = 8
)
```

This model estimates the age composition across the specified age range and
links age to length using the aged fish in `CA`.

A more flexible model can include additional structure, for example random year
effects or spatial smooths:

```{r eval = FALSE}
ALK_spatial <- fitALK(
  dab_Q1,
  minAge = 2,
  maxAge = 8,
  model = "cra ~ LngtCm + s(Year, bs = 're') + s(lon, lat, bs = 'ts')"
)
```

Such extensions can be useful when the age-length relationship varies over time
or space, or when sampling is spatially structured.

## How the stochastic ALK is formulated

`fitALK()` uses a continuation-ratio formulation. For each age threshold `a`,
the data are restricted to fish with `Age >= a`, and a binary model is fitted
for whether `Age > a`. In other words, for each threshold it estimates

\[
P(\mathrm{Age} > a \mid \mathrm{Age} \ge a,\; x)
\]

where `x` denotes the covariates, such as length, year, or location.

The full ALK is then constructed by combining these conditional probabilities
across ages. This formulation has two practical advantages:

- it guarantees that predicted probabilities across age classes sum to one, and
- it allows flexible modelling of covariate effects on the age distribution.

# Predict catch at age from the fitted ALK

Once the ALK has been fitted, it can be used to convert numbers at length into
numbers at age. Prediction is currently done using the corresponding prediction
method from the **DATRAS** package:

```{r}
dab_Q1$Nage <- predict(ALK)
```

This adds a new matrix to the `HH` component with the predicted numbers by age:

```{r}
head(dab_Q1$Nage)
```

Each row corresponds to a haul, and each column corresponds to an age class.
The values are derived by combining the representative numbers at length from
`HL` with the estimated probabilities from the ALK.

These haul-level age compositions can then be used for a range of downstream
analyses. For example, they can be aggregated to produce age-based survey
indices, compared among years or areas, or used as input to stock-assessment
models.

## Quick inspection of predicted catch at age

A simple first check is to look at the total predicted numbers by age across all
hauls:

```{r}
colSums(dab_Q1$Nage, na.rm = TRUE)
```

This gives the overall predicted age composition for the selected subset of the
survey. If desired, age-specific maps or time-series summaries can be derived in
exactly the same way as for other haul-level quantities stored in `HH`.

# Summary

Length-at-age data provide the link between size-based survey observations and
age-based population dynamics, and are therefore central to many fisheries
applications. Within the broader **DATRASextra** workflow, this vignette builds
on the data access, cleaning, and length-based summaries introduced in the
earlier vignettes, and extends them toward age-structured analyses. In
particular, stochastic age-length keys make it possible to combine the
catch-representative length information in `HL` with the aged subsample in `CA`,
thereby deriving age-based summaries that are suitable for stock assessment,
survey standardisation, and analyses of temporal or spatial population
structure.

At the same time, the resulting age compositions are only as reliable as the
underlying age data and the assumptions of the ALK model. Age samples may be
sparse, unevenly distributed in space or time, or unrepresentative of the full
length range, and the age-length relationship itself may vary among years,
areas, or survey strata. For this reason, ALKs should not be treated as a purely
technical conversion step, but as a statistical model whose assumptions and
predictive behaviour deserve careful checking. In practice, this often means
comparing alternative model formulations, examining residual patterns, and
considering whether additional covariates such as year, quarter, sex, or space
should be included.

This vignette therefore provides a starting point rather than an endpoint. Once
age-based haul summaries have been derived, they can be aggregated into survey
indices, used to study shifts in recruitment or age structure, or linked with
the other **DATRASextra** tools for biomass, length composition, and spatial
analyses. A natural next step is to compare age-based and length-based views of
the survey data, evaluate the stability of ALKs over time, and assess how the
choice of ALK model influences downstream scientific conclusions and fisheries
advice.

# References
