---
title: "How to recreate the FishGlob DATRAS dataset"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```


## Introduction

This vignette demonstrates how to recreate the **DATRAS** part of the **FishGlob dataset** using functions from **DATRASextra**.

> **Important:**  
> This workflow requires the **full raw DATRAS HH/HL/CA dataset** stored locally.  
> The examples assume your raw data are under: `temporary/data/`.

This is how you can download the dataset:
```{r, eval=FALSE}
surveys <- c(
  "NS-IBTS",
  "BITS",
  "EVHOE",
  "FR-CGFS",
  "IE-IGFS",
  "NIGFS",
  "SCOROC",
  "SWC-IBTS",
  "SCOWCGFS",
  "SP-PORC"
)

dir <- here('temporary/data')
years <- c(1993:2025)

downloadDATRAS(surveys = surveys, years = years, dir = dir)
```


## Load libraries

```{r}
library(DATRASextra)
library(here)
library(tidyverse) # sorry Tobi :P
```

## Read data

```{r}
raw <- readDATRAS(here("temporary/data")) %>%
  suppressWarnings()
#str(raw)
names(raw)
```

## Cleaning and pruning

```{r}
dat <- cleanFishglob(raw)
```

Reduce object size removing columns not needed

```{r}
object.size(dat) %>% format(units='Mb')
dat <- pruneFishglob(dat)
object.size(dat) %>% format(units='Mb')
```

## Add swept area (and impute when missing)

```{r}
dat <- addSweptAreaFishGlob(dat)
```

## Add biomass estimates (length-weight)

```{r}
dat <- addWeightFishglob(dat)
```

## Format into FishGlob output structure

```{r}
datras <- formatFishglob.DATRASraw(dat)
head(datras)
```

## Compare it with FishGlob

### Download FishGlob
```{r}
# load survey data --------------------------------------------------------
fishglob_url <- "https://github.com/AquaAuma/FishGlob_data/raw/d71dfa03c2912b4e9d9cd10412ae2af52ba56ae5/outputs/Compiled_data/FishGlob_public_std_clean.RData" # stick with this version
options(timeout = 300) # increase timeout to 5 minutes
load(url(fishglob_url))
fishglob <- data
```

### Get the surveys in common
```{r}

common_surveys <- intersect(fishglob$survey, datras$survey)

fishglob_common <- fishglob[fishglob$survey %in% common_surveys, ]
datras_common <- datras[datras$survey %in% common_surveys, ]

fishglob_common$haul_id <- gsub(" ", ":", fishglob_common$haul_id)

# get common haul ids
common_haul_ids <- intersect(fishglob_common$haul_id, datras_common$haul_id)

# fishglob
fishglob_agg <- fishglob_common %>% filter(haul_id %in% common_haul_ids) %>% 
  group_by(haul_id, accepted_name, survey) %>%
  summarise(wgt_cpua = sum(wgt_cpua, na.rm = TRUE), num = sum(num, na.rm = TRUE), .groups = "drop")

# datras_extra
datras_agg <- datras_common %>% filter(haul_id %in% common_haul_ids) %>% 
  group_by(haul_id, accepted_name, survey) %>%
  summarise(wgt_cpua = sum(wgt_cpua, na.rm = TRUE), num = sum(num, na.rm = TRUE), .groups = "drop")

merged_agg <- merge(
  fishglob_agg,
  datras_agg,
  by = c("survey","haul_id", "accepted_name"),
  suffixes = c("_fishglob", "_datras")
)

df_cod <- merged_agg %>% 
  filter(accepted_name == "Gadus morhua") # compare cod as example

```

```{r setup2, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.path = "vignettes/articles/figures/",  # all figures go here
  fig.width = 9,
  fig.height = 7,
  dpi = 450
)
```


```{r plot_num, fig.width=9, fig.height=7}
ggplot(df_cod, aes(x = num_fishglob, 
                   y = num_datras)) +
  
  # Points
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0,
              color = "red", linewidth = 1) +
  labs(
    x = "FishGlob numbers",
    y = "DATRASextra numbers",
    title = "Comparison of num per haul by survey for cod"
  ) +
  facet_wrap(~ survey, scales = "free")

```

```{r plot_wg, fig.width=9, fig.height=7}

ggplot(df_cod, aes(x = wgt_cpua_fishglob, 
                   y = wgt_cpua_datras)) +
  
  # Points
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0,
              color = "red", linewidth = 1) +
  labs(
    x = "FishGlob kg/km2",
    y = "DATRASextra kg/km2",
    title = "Comparison of wgt_cpua per haul by survey for cod"
  ) +
  
  facet_wrap(~ survey, scales = "free")


```