---
title: "Extracting Length-at-Age (LAA) from DATRAS"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```


## Goal

This vignette demonstrates how to extract and explore length-at-age (LAA) data
from a cleaned DATRAS object and how to estimate stochastic age-length keys
(ALKs).

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(DATRASextra)
```

This vignette assumes you already have a cleaned DATRAS object (see the tutorial
vignette on data preparation). Here we use the example data:

```{r}
data("dab")
```

A typical DATRAS object is a list with three core data sets:

```{r}
str(dab, max.level = 1)
```

- `CA`: individual-level biological data (e.g., length, weight, sex, maturity, age)
- `HH`: haul/station-level metadata (e.g., position, time, gear, environment)
- `HL`: catch-at-length and subsampling information

When a species appears in `CA`, a corresponding `HL` record should also exist
(even if values are missing codes such as `-9`).

---

## Select years and clean_datras data

For LAA analyses, `CA` is essential because it contains age readings. However,
`CA` is not catch-representative by itself, so ALKs are needed to convert
catch-at-length (`HL`) into catch-at-age.

We first select years of interest:

```{r}
years <- 2020:2024
```

Then apply a quick standard cleaning/filtering step:

```{r}
dab <- clean_datras(dab, years = years)
```

Equivalent manual filtering (shown for reference):

```{r eval=FALSE}
dab <- subset(
  dab,
  Year %in% years &
    HaulVal == "V" &
    DayNight == "D" &
    StdSpecRecCode == 1
)
```

---

## Investigate length-at-age data

```{r}
ca <- dab[["CA"]]
str(ca, max.level = 1)
```

### Sample size by year

```{r}
# Number of individual rows per Year
xtabs(~ Year, ca)

# Number of unique hauls per Year
xtabs(~ Year, data = unique(ca[c("Year", "haul.id")]))
```

### Add spatial information from HH

Some variables (e.g., haul position) are in `HH` rather than `CA`. Merge them by
`haul.id`:

```{r}
ca <- merge(
  ca,
  dab[["HH"]][c("haul.id", "lon", "lat")],
  by = "haul.id",
  all.x = TRUE,
  sort = FALSE,
  suffixes = c("", ".HH")
)
```

Now you can inspect spatial patterns in LAA observations:

```{r fig.width=6, fig.height=5}
plot(ca$lon, ca$lat, cex = pmax(0.6, ca$Age / 3), xlab = "Longitude", ylab = "Latitude")
```

### Age availability and missingness

```{r}
table(ca$Age, useNA = "ifany")
```

```{r}
sum(is.na(ca$Age))
```

Around `r round(mean(is.na(ca$Age)) * 100, 1)`% of `CA` entries have missing age.

For ALK estimation, `NoAtALK` is the frequency/count associated with each
age-length observation.

---

## Estimate age-length keys (ALKs)

First, add the length spectrum from `HL`:

```{r}
dab <- addSpectrum(dab)
```

ALKs are often estimated by quarter for stock-assessment workflows. Example for
Q1:

```{r}
dab_Q1 <- subset(dab, Quarter == 1)
```

Fit a basic stochastic ALK:

```{r}
ALK <- fitALK(
  dab_Q1,
  minAge = 3, ## TODO
  maxAge = 8
)
```

A more flexible model can include random year effects and spatial smooths:

```{r eval = FALSE}
ALK_spatial <- fitALK(
  dab_Q1,
  minAge = 2,
  maxAge = 8,
  model = "cra ~ LngtCm + s(Year, bs = 're') + s(lon, lat, bs = 'ts')"
)
```

### Model interpretation

`fitALK()` uses a continuation-ratio formulation. For each age threshold `a`,
the data are restricted to fish with `Age >= a`, and the binary response is
`Age > a`. This estimates:

\[
P(\text{Age} > a \mid \text{Age} \ge a,\; x)
\]

Combining these conditional probabilities across ages yields the full ALK.

---

## Predict from the fitted ALK

Exact prediction workflows depend on your package version and chosen downstream
pipeline. A typical pattern is:

```{r eval=FALSE}
# Example pattern (adapt to your object structure)
newdata <- dab_Q1[["HL"]]
pred <- predict(ALK_spatial, newdata = newdata, type = "response")

# Continue with conversion from catch-at-length to catch-at-age per haul
# using the predicted age probabilities.
```

If your workflow uses a package-specific helper (e.g., an ALK application
function), use that helper in place of `predict()`.

---

## Summary

In this vignette, we:

1. Loaded and cleaned DATRAS data
2. Explored individual age data in `CA`
3. Added haul-level spatial context from `HH`
4. Estimated stochastic ALKs (basic and spatially enriched)
5. Outlined how to predict/apply ALKs to convert catch-at-length to catch-at-age
