---
title: "Building a spatiotemporal prediction grid"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Building a spatiotemporal prediction grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width = 6,
                      fig.height = 5,
                      dev = "png",
                      dev.args = list(bg = "white"))

knitr::knit_hooks$set(
  plot = function(x, options) {
    paste0(
      "<div style='background:white; padding:0.5rem; display:inline-block;'>",
      knitr::hook_plot_html(x, options),
      "</div>"
    )
  }
)

oldpar <- par(no.readonly = TRUE)
par(bg = "white")
```

This article shows how to build a regular prediction grid over a survey domain
with `make_survey_grid()`, and how to extend it into a *spatiotemporal* grid by
repeating it across years. Such a grid is a common input for spatial and
spatiotemporal species distribution models, where predictions are made on a
regular lattice covering the surveyed area.

We use the example data set `dab` (dab, *Limanda limanda*, from the
[NS-IBTS](https://gis.ices.dk/geonetwork/srv/eng/catalog.search#/metadata/abe375f9-fcf0-4839-a842-1f28650c98d3)
survey, 2020-2023).

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Load the packages
library(DATRASextra)
library(sf)

## Load the example data
data(dab)
```


# Disclaimer

This is just one way of building a prediction grid. Choices such as the
resolution, the distance threshold, and whether to remove land are
subjective and should be tailored to your own data and objectives.


# Trawl locations

We start from the unique haul positions in the `HH` table. `make_survey_grid()`
works directly on coordinate vectors, so we only need the `lon` and `lat`
columns.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Unique trawl locations
trawls <- unique(dab[["HH"]][, c("haul.id", "lon", "lat")])
head(trawls)
```


# A spatial grid

`make_survey_grid()` builds an equally spaced grid covering the range of the
coordinates. `resolution` and `max_dist` are expressed in the **same units** as
the coordinates — here degrees, because the hauls are in lon/lat. We use a
0.1° grid and drop grid nodes that are more than 0.3° from any haul, so the grid
stays close to where the survey actually sampled (this needs the `RANN`
package).

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Regular grid near the trawl observations
grid <- make_survey_grid(
  x          = trawls$lon,
  y          = trawls$lat,
  resolution = 0.1,
  max_dist   = 0.3
)

nrow(grid)
head(grid)
```

The result is a data frame with columns `X` and `Y`. A quick plot shows the
grid following the survey footprint:

```{r, eval = TRUE, echo = TRUE, fig.alt = "Prediction grid nodes with trawl observations overlaid"}
## Grid nodes (blue) and trawl observations (red)
plot(grid$X, grid$Y, pch = 15, col = "lightblue", cex = 0.5,
     xlab = "Longitude", ylab = "Latitude")
points(trawls$lon, trawls$lat, col = "red", pch = 20, cex = 0.6)
legend("topright", legend = c("Prediction grid", "Trawl observations"),
       col = c("lightblue", "red"), pch = c(15, 20), bg = "white")
box(lwd = 1.5)
```

For predictions in true distance units, project the coordinates first (for
example to UTM), pass the projected coordinates to `make_survey_grid()`, and set
`resolution` and `max_dist` in metres or kilometres.


# Adding the time dimension

To turn the spatial grid into a *spatiotemporal* one, pass a vector of years to
`time`. The spatial grid is then repeated for each year and a `year` column is
added, giving one row per grid node and year.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Repeat the grid across survey years
grid_st <- make_survey_grid(
  x          = trawls$lon,
  y          = trawls$lat,
  resolution = 0.1,
  max_dist   = 0.2,
  time       = 2020:2023
)

nrow(grid_st)
head(grid_st)
```

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## One spatial grid per year
table(grid_st$year)
```

This `grid_st` data frame - with `X`, `Y`, and `year` - can be passed directly
to a spatiotemporal model as the set of locations to predict on.


# Projected coordinates

Distances in degrees are not equal in all directions: in the North Sea a
degree of longitude is only about 60 % of a degree of latitude. For a grid in
kilometres, project the haul positions first. `add_xy()` adds the columns `X`
and `Y` (by default the European equal-area projection EPSG:3035, in km) to
`HH`, and with `inverse = TRUE` adds longitude and latitude to a projected
grid.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Haul positions in km
dab_xy <- add_xy(dab)
hauls <- unique(dab_xy[["HH"]][, c("haul.id", "X", "Y", "Year", "Quarter", "Depth")])
hauls$Year <- as.numeric(as.character(hauls$Year))

## 10 km grid, nodes within 15 km of a haul
grid_km <- make_survey_grid(hauls$X, hauls$Y, resolution = 10, max_dist = 15)

## Longitude and latitude of each node
grid_km <- add_xy(grid_km, inverse = TRUE)
head(grid_km)
```


# Filling gaps

With `max_dist` alone, a grid node is kept only when a haul lies within
`max_dist`. Where the hauls are farther apart than `2 * max_dist`, this
leaves holes inside the survey area and notches along its edge, although the
survey covers the area between them. A larger `max_dist` would close the
holes, but would also push the outer edge further from the outermost hauls.

`fill_gaps` closes holes and bays narrower than about `2 * fill_gaps` and
leaves the outer edge where `max_dist` put it. The footprint is first grown by
`max_dist + fill_gaps` around each haul, then shrunk by `fill_gaps` again.

```{r, eval = TRUE, echo = TRUE, fig.width = 9, fig.height = 4.5, fig.alt = "Prediction grid without and with gap filling"}
grid_filled <- make_survey_grid(hauls$X, hauls$Y, resolution = 10,
                                max_dist = 15, fill_gaps = 20)
c(without = nrow(grid_km), with = nrow(grid_filled))

grids <- list("max_dist = 15" = grid_km,
              "max_dist = 15, fill_gaps = 20" = grid_filled)
par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
for (nm in names(grids)) {
  g <- grids[[nm]]
  plot(g$X, g$Y, pch = 15, cex = 0.4, col = "lightblue", asp = 1,
       xlab = "", ylab = "", main = nm)
  points(hauls$X, hauls$Y, pch = 20, cex = 0.3, col = "red")
}
```


# Where the survey sampled each year

A spatiotemporal model predicts into every node in every year, also in years
in which no haul came near a node. `add_grid_support()` flags this. It adds
`supported`, which is `TRUE` when the hauls of a year, or of the years at most
`window` away, cover the node, and `coverage`, the share of years in which a
node is covered. Coverage is checked with the same `max_dist` and `fill_gaps`
as the grid.

Surveys often cover different areas in different quarters, so the grid is
built per quarter, here for quarter 1.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
years <- sort(unique(hauls$Year))
q1 <- hauls[hauls$Quarter == "1", ]

grid_q1 <- make_survey_grid(hauls$X, hauls$Y, resolution = 10, max_dist = 15,
                            fill_gaps = 20, time = years)
grid_q1 <- add_grid_support(grid_q1, q1$X, q1$Y, q1$Year,
                            max_dist = 15, fill_gaps = 20, window = 1)

## Keep the nodes covered in at least a quarter of the years
grid_q1 <- grid_q1[grid_q1$coverage >= 0.25, ]

## Share of supported nodes per year
tapply(grid_q1$supported, grid_q1$year, mean)
```

Predictions for unsupported nodes are extrapolations. They can be dropped,
or kept and marked on the maps.


# Adding depth

Depth is a common covariate in distribution models, and a check that the grid
stays within the depths the survey fishes. `add_bathymetry()` adds the depth
from NOAA's ETOPO bathymetry (through the `marmap` package). With `path`, the
download is kept and reused. `depth_range` adds `depth_ok`, whether a node lies
within a depth range, here that of 98 % of the hauls.

```{r, eval = FALSE, echo = TRUE}
grid_q1 <- add_bathymetry(
  grid_q1,
  path = tempdir(),
  depth_range = quantile(q1$Depth, c(0.01, 0.99), na.rm = TRUE)
)

## Drop land, which has no depth
grid_q1 <- grid_q1[!is.na(grid_q1$Depth), ]
table(grid_q1$depth_ok)
```


# Removing land (optional)

A rectangular grid can include cells that fall on land. If your model domain is
sea only, drop those cells. Here we use coastline polygons from Natural Earth
and keep only grid nodes that fall in the sea.

```{r, eval = FALSE, echo = TRUE, fig.alt = ""}
library(rnaturalearth)

## Land polygon (WGS84)
land <- st_union(st_make_valid(ne_countries(scale = 10, returnclass = "sf")))

## Keep only grid nodes that are not on land
grid_sf <- st_as_sf(grid, coords = c("X", "Y"), crs = 4326)
on_land <- lengths(st_intersects(grid_sf, land)) > 0
grid_sea <- grid[!on_land, ]
```

Cross `grid_sea` with the survey years (as above) to obtain a sea-only
spatiotemporal grid.

```{r, echo=FALSE}
par(oldpar)
```
