---
title: "Building a spatiotemporal prediction grid"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Building a spatiotemporal prediction grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width = 6,
                      fig.height = 5,
                      dev = "png",
                      dev.args = list(bg = "white"))

knitr::knit_hooks$set(
  plot = function(x, options) {
    paste0(
      "<div style='background:white; padding:0.5rem; display:inline-block;'>",
      knitr::hook_plot_html(x, options),
      "</div>"
    )
  }
)

oldpar <- par(no.readonly = TRUE)
par(bg = "white")
```

This article shows how to build a regular prediction grid over a survey domain
with `make_survey_grid()`, and how to extend it into a *spatiotemporal* grid by
repeating it across years. Such a grid is a common input for spatial and
spatiotemporal species distribution models, where predictions are made on a
regular lattice covering the surveyed area.

We use the example data set `dab` (dab, *Limanda limanda*, from the
[NS-IBTS](https://gis.ices.dk/geonetwork/srv/eng/catalog.search#/metadata/abe375f9-fcf0-4839-a842-1f28650c98d3)
survey, 2020-2023).

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Load the packages
library(DATRASextra)
library(sf)

## Load the example data
data(dab)
```


# Disclaimer

This is just one way of building a prediction grid. Choices such as the
resolution, the distance threshold, and whether to remove land are
subjective and should be tailored to your own data and objectives.


# Trawl locations

We start from the unique haul positions in the `HH` table. `make_survey_grid()`
works directly on coordinate vectors, so we only need the `lon` and `lat`
columns.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Unique trawl locations
trawls <- unique(dab[["HH"]][, c("haul.id", "lon", "lat")])
head(trawls)
```


# A spatial grid

`make_survey_grid()` builds an equally spaced grid covering the range of the
coordinates. `resolution` and `max_dist` are expressed in the **same units** as
the coordinates — here degrees, because the hauls are in lon/lat. We use a
0.1° grid and drop grid nodes that are more than 0.3° from any haul, so the grid
stays close to where the survey actually sampled (this needs the `RANN`
package).

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Regular grid near the trawl observations
grid <- make_survey_grid(
  x          = trawls$lon,
  y          = trawls$lat,
  resolution = 0.1,
  max_dist   = 0.3
)

nrow(grid)
head(grid)
```

The result is a data frame with columns `X` and `Y`. A quick plot shows the
grid following the survey footprint:

```{r, eval = TRUE, echo = TRUE, fig.alt = "Prediction grid nodes with trawl observations overlaid"}
## Grid nodes (blue) and trawl observations (red)
plot(grid$X, grid$Y, pch = 15, col = "lightblue", cex = 0.5,
     xlab = "Longitude", ylab = "Latitude")
points(trawls$lon, trawls$lat, col = "red", pch = 20, cex = 0.6)
legend("topright", legend = c("Prediction grid", "Trawl observations"),
       col = c("lightblue", "red"), pch = c(15, 20), bg = "white")
box(lwd = 1.5)
```

For predictions in true distance units, project the coordinates first (for
example to UTM), pass the projected coordinates to `make_survey_grid()`, and set
`resolution` and `max_dist` in metres or kilometres.


# Adding the time dimension

To turn the spatial grid into a *spatiotemporal* one, pass a vector of years to
`time`. The spatial grid is then repeated for each year and a `year` column is
added, giving one row per grid node and year.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Repeat the grid across survey years
grid_st <- make_survey_grid(
  x          = trawls$lon,
  y          = trawls$lat,
  resolution = 0.1,
  max_dist   = 0.2,
  time       = 2020:2023
)

nrow(grid_st)
head(grid_st)
```

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## One spatial grid per year
table(grid_st$year)
```

This `grid_st` data frame - with `X`, `Y`, and `year` - can be passed directly
to a spatiotemporal model as the set of locations to predict on.


# Removing land (optional)

A rectangular grid can include cells that fall on land. If your model domain is
sea only, drop those cells. Here we use coastline polygons from Natural Earth
and keep only grid nodes that fall in the sea.

```{r, eval = FALSE, echo = TRUE, fig.alt = ""}
library(rnaturalearth)

## Land polygon (WGS84)
land <- st_union(st_make_valid(ne_countries(scale = 10, returnclass = "sf")))

## Keep only grid nodes that are not on land
grid_sf <- st_as_sf(grid, coords = c("X", "Y"), crs = 4326)
on_land <- lengths(st_intersects(grid_sf, land)) > 0
grid_sea <- grid[!on_land, ]
```

Cross `grid_sea` with the survey years (as above) to obtain a sea-only
spatiotemporal grid.

```{r, echo=FALSE}
par(oldpar)
```
