---
title: "Matching EMODnet seabed habitats to hauls"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Matching EMODnet seabed habitats to hauls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7
)
```

This article shows how to attach broad-scale seabed habitat information from
[EMODnet Seabed Habitats](https://emodnet.ec.europa.eu/en/seabed-habitats)
(the EUSeaMap product) to the `HH` (haul) table of a `datras_raw` object. The
habitat map is a polygon (vector) layer, so the haul positions in `HH` are
turned into points and matched into the habitat polygons with a spatial join.

EUSeaMap is a large, continental data set. Downloading the polygons for a whole
survey at once is slow, so we work with a small set of hauls in one area and
download only the habitat polygons that cover them. The same code scales to a
full survey if you are prepared for a larger download (see the note at the end).

We use the example data set `dab` and the **sf** package.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Load the packages
library(DATRASextra)
library(sf)

## EMODnet habitat polygons are not always valid under sf's spherical (S2)
## geometry engine, which makes spatial joins fail. Use planar geometry, which
## is appropriate for a regional layer like this.
sf_use_s2(FALSE)

## Load the example data
data(dab)
```


# Haul positions as points

To keep the habitat download small, we first restrict the object to the hauls
from one small area (a box in the central North Sea) with `subset()`. Working on
this subset throughout means the habitat we attach later lines up with every
haul in the object, with no unmatched (`NA`) hauls from areas we did not
download.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Restrict to a small area (central North Sea)
box <- subset(dab, lon > 2 & lon < 4 & lat > 54 & lat < 55)
nrow(box[["HH"]])
```

The `HH` table stores haul positions in the `lon` / `lat` columns
(longitude/latitude, WGS84). We build an `sf` point layer from them, keeping
`haul.id` so the habitat can be matched back to `HH` afterwards.

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Haul points (EPSG:4326 = WGS84 lon/lat)
hauls <- st_as_sf(
  box[["HH"]][, c("haul.id", "lon", "lat")],
  coords = c("lon", "lat"),
  crs = 4326
)

hauls
```


# Download the EUSeaMap layer

EUSeaMap is served through the EMODnet Seabed Habitats WFS. Requesting the whole
European map would be very large, so we restrict the download to the bounding
box of the hauls.

First, the haul extent (in lon/lat):

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Bounding box of the hauls
bb <- st_bbox(hauls)
bb
```

The EMODnet Seabed Habitats WFS endpoint is
`https://ows.emodnet-seabedhabitats.eu/geoserver/emodnet_open/ows`. You can list
the available layers and their attributes from the service capabilities:

```{r, eval = FALSE, echo = TRUE, fig.alt = ""}
## Discover available layers (run once, interactively)
#wfs <- "https://ows.emodnet-seabedhabitats.eu/geoserver/emodnet_open/ows"
#st_layers(paste0("WFS:", wfs))
```

Pick the EUSeaMap layer you need. The European broad-scale map classified with
EUNIS 2019 is `emodnet_open:eusm2025_eunis2019_full`. Download only the haul
bounding box by passing a `BBOX` filter in the WFS `GetFeature` request. The
request below asks for GeoJSON within the haul extent, and `st_make_valid()`
repairs the polygons so they can be used in a spatial join:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## WFS GetFeature request, restricted to the haul bounding box
wfs   <- "https://ows.emodnet-seabedhabitats.eu/geoserver/emodnet_open/ows"
layer <- "emodnet_open:eusm2025_eunis2019_full"

url <- paste0(
  wfs,
  "?service=WFS&version=2.0.0&request=GetFeature",
  "&typeNames=", layer,
  "&srsName=EPSG:4326",
  "&bbox=", paste(bb["ymin"], bb["xmin"], bb["ymax"], bb["xmax"], "urn:ogc:def:crs:EPSG::4326", sep = ","),
  "&outputFormat=application/json"
)

habitat <- st_read(url, quiet = TRUE)
habitat <- st_make_valid(habitat)

## Inspect the attributes and choose the habitat column you want to keep
names(habitat)
```

The key habitat attributes in this layer are `eunis2019c` (EUNIS 2019 code),
`eunis2019d` (EUNIS 2019 description), and `all2019d` (full habitat
description), alongside the component descriptors `substrate`, `energy`,
`biozone`, `salinity`, and `oxygen`.

Note that WFS axis order for EPSG:4326 is *lat,lon*, which is why the `bbox`
string above is ordered `ymin,xmin,ymax,xmax`. If you prefer, you can instead
download the layer once from the
[EMODnet Seabed Habitats download portal](https://emodnet.ec.europa.eu/geoviewer/)
and read the local file with `st_read()`.


# Spatial join to the hauls

With the habitat polygons in memory, transform the haul points to the habitat
layer's CRS and join each point to the polygon it falls in. Keep only the
habitat attribute(s) you want (here the seabed `substrate`; use `eunis2019d` for
the full EUNIS habitat description):

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Name of the habitat attribute to keep (from names(habitat) above)
habitat_col <- "substrate"

## Match points to polygons in the habitat layer's CRS
hauls_m <- st_transform(hauls, st_crs(habitat))
joined  <- st_join(hauls_m, habitat[habitat_col], join = st_intersects)

## One habitat value per haul.id
head(st_drop_geometry(joined))
```

Because polygon boundaries can overlap or hauls can sit on a border, make sure
there is exactly one row per haul before writing back (keep the first match):

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Collapse to one row per haul.id
joined <- st_drop_geometry(joined)
joined <- joined[!duplicated(joined$haul.id), ]
```


# Attach the habitat to HH

Finally, match the habitat back onto `HH` by `haul.id` and store it as a new
column. Matching by `haul.id` (rather than row order) is robust to any reordering
during the join:

```{r, eval = TRUE, echo = TRUE, fig.alt = ""}
## Add the habitat column to HH
box[["HH"]]$habitat <- joined[[habitat_col]][
  match(box[["HH"]]$haul.id, joined$haul.id)
]

## Check the result
table(box[["HH"]]$habitat, useNA = "ifany")
```

Because we worked on the `box` subset throughout, every haul in it gets a
habitat (apart from any haul that genuinely falls outside the mapped area, which
stays `NA`). The `HH` table now carries a habitat class for each haul, which can
be used as a covariate in downstream analyses or as a grouping variable when
summarising catch by habitat.


# Map the matched hauls

Finally, a quick base-R map of the selected hauls coloured by their newly
matched substrate. We turn the `HH` rows back into an `sf` point layer and let
the `sf` plot method colour the points by the `habitat` column:

```{r, eval = TRUE, echo = TRUE, fig.alt = "Map of the selected hauls coloured by their matched EMODnet seabed substrate"}
## Selected hauls as points, carrying the matched substrate
hauls_hab <- st_as_sf(
  box[["HH"]][, c("haul.id", "lon", "lat", "habitat")],
  coords = c("lon", "lat"),
  crs = 4326
)

## Colour each haul by its matched substrate
plot(hauls_hab["habitat"], pch = 19,
     main = "Selected hauls coloured by EMODnet seabed substrate")
```

# Summary

Matching EMODnet seabed habitats to a `datras_raw` object comes down to three
steps: turn `HH$lon` / `HH$lat` into an `sf` point layer, download the EUSeaMap
polygons for the haul bounding box from the EMODnet Seabed Habitats WFS, and
join the points into the polygons with `st_join()`. The same pattern works for
other EMODnet vector layers by changing the WFS layer name.

For a whole survey the bounding box can be large and the single WFS download
slow. In that case, split the hauls into smaller boxes (for example by ICES
rectangle or a coarse lon/lat grid), download and join each box separately, and
combine the results — or download the EUSeaMap layer once from the
[download portal](https://emodnet.ec.europa.eu/geoviewer/) and read it locally.
