---
title: "Extract numbers or biomass for a single species"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Extract numbers or biomass for a single species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  eval = FALSE,
  error = FALSE,
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7
)
```

```{r setup-example}
library(DATRASextra)

# select survey
survey <- "NS-IBTS"

# create temporary directory
tmp <- file.path(tempdir(), "DATRASextra")
dir.create(tmp, recursive = TRUE, showWarnings = FALSE)

# download data
download_datras(path = tmp, surveys = survey, years = 2021)

# read in
surv0 <- read_datras(file.path(tmp, survey))

# clean data
surv <- clean_datras(surv0)

# subset for your species
surv <- subset(surv, Valid_Aphia == "126822") # Horse mackerel
```

To calculate total numbers and weight by haul, the information in the `HL`
table first needs to be raised to the haul level. A convenient first step is
to add numbers-at-length:

```{r}
surv <- add_numbers_at_length(surv)
```

This adds a matrix `N` to the `HH` table, containing the estimated numbers in
each length class for each haul. The length classes can be modified with the
`cm_breaks` or `by` arguments.

The total number by haul can then be obtained by summing over length classes:

```{r}
surv <- add_total_numbers_by_haul(surv)
```

By default, this adds a single column `HaulN` to the `HH` table. More generally,
custom bins can be specified via `length_cuts`. In that case, the result may
contain separate totals for broader user-defined length groups.

If species-specific length and weight information are available in `CA`, or
suitable length-weight relationships are available in the `species_info` table,
the numbers can also be converted to weight:

```{r}
surv <- add_total_weight_by_haul(surv)
```

This adds `HaulWgt` to the `HH` table, containing the total estimated weight by
haul. As for abundance, custom length-group summaries can be generated when
`length_cuts` is used.
