---
title: "Plotting DATRAS overviews"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Plotting DATRAS overviews}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7
)
```

## Overview

This vignette demonstrates the unified plotting function
`plot_datras_overview()`. The function allows to generate a quick overview of
all haul locations in DATRAS or number of hauls per ICES Statistical Rectangle,
but also allows to generate more complicated maps of any variables (and offset
variable) by haul location in a point plot or gridded plot of a specific data
set. Specifically, the function supports:

- point maps (`mode = "points"`)
- gridded maps (`mode = "grid"`)
- multiple grid metrics (`presence`, `count_hauls`, `count_surveys`, `sum`, `mean`)
- grouping and faceting (`by_survey`, `by_gear`, `by_quarter`, `multi_panels`)
- raw or StatRec spatial basis (`spatial_basis = "raw"` or `"statrec"`)
- value/offset/transform workflows for quantitative overlays

Load the package with:

```{r setup}
library(DATRASextra)
```


## Quick Start

If no object is supplied (`x = NULL`), the function uses package survey overview
data, which contains all surveys and hauls in DATRAS:

```{r quick-null}
plot_datras_overview()
```

To get an overview over the surveys:

```{r quick-null-surv}
plot_datras_overview(by_survey = TRUE)
```

If the legend is in the way, the multiple different control arguments can be
used to place and modify the legend:

```{r quick-null-surv-legend}
plot_datras_overview(by_survey = TRUE,
                     legend_ncol = 6,
                     legend_pos = "bottom",
                     legend_cex = 0.8)
```

With so many surveys and repeating colours, it might be easier to plot the
surveys in separate panels:

```{r null-surv}
plot_datras_overview(by_survey = TRUE,
                     multi_panels = TRUE)

```

Besides the general overview over all surveys, the function also allows to
create a visual overview of a specific data set, using for example the `mini`
data set of `DATRASextra`:

```{r quick-mini-points}
plot_datras_overview(mini)
```


## Aggregate by

Besides surveys, the function allows to quickly create a comparison between
gears:

```{r agg-gear}
plot_datras_overview(mini, by_gear = TRUE)
```

by quarter:

```{r agg-quarter}
plot_datras_overview(mini, by_quarter = TRUE)
```

by year:

```{r agg-year}
plot_datras_overview(mini, by_year = TRUE)
```

or day and night:

```{r agg-dn}
plot_datras_overview(mini, by_daynight = TRUE)
```

or any combination of them:

```{r agg-gear-survey}
plot_datras_overview(mini,
                     by_gear = TRUE,
                     by_survey = TRUE)
```

setting the `multi_panel` argument to `TRUE` avoids the overlap and might help interpretation:

```{r agg-gear-survey-multi}
plot_datras_overview(mini, by_gear = TRUE,
                     by_survey = TRUE,
                     multi_panel = TRUE)
```


## Points vs. gridded

By default, the function plots the actual haul locations, but it might be
preferrable to aggregate the hauls and plot them by ICES statistical rectangle
midpoints by setting the argument `spatial_basis = "statrec"`:

```{r spat_basis}
plot_datras_overview(mini,
                     by_gear = TRUE,
                     spatial_basis = "statrec")
```

As multiple levels (here gears) might be present in a single ICES statistical
rectangle, by default, the dominant level is assigned to that rectangle, but the
argument `grid_group_strategy` allows to change that behaviour.

Instead of plots by statistical rectangle, the function also allows to plot an
gridded image plot. This can be done by setting the `mode = "grid"`:

```{r mode-grid}
plot_datras_overview(mini, by_gear = TRUE, mode = "grid")
```

Again, by default the dominant level is shown for each grid cell.



## Plotted quantity

While so far, the plots only quantified the absence / presence of a haul with a
specific charactistics in each area, the function also allows us to plot various
quantities by using the `metric` argument. We can for example plot the number of
hauls with:

```{r metric-count-hauls}
plot_datras_overview(mini,
                     mode = "grid",
                     metric = "count_hauls")
```

Note that this only works for the gridded mode and if requires the `multi_panels
= TRUE` if you want to split it by another variable:

```{r metric-count-hauls-y}
plot_datras_overview(mini,
                     mode = "grid",
                     metric = "count_hauls",
                     by_year = TRUE,
                     multi_panels = TRUE)
```

Similarly, you can plot the number of surveys by quarter:

```{r metric-count-surv-q}
plot_datras_overview(mini,
                     mode = "grid",
                     metric = "count_surveys",
                     by_quarter = TRUE,
                     multi_panels = TRUE)
```

Other options of the `metric` argument are `"sum"` or `"mean"`, but they become
more meaningful when combined with the `value_var` argument (see below).



## Value / Offset / Transform Workflows

If your DATRAS data set includes quantitative columns (for example a response
and an effort-like offset), you can map transformed values in both grid and
point modes. For example, a quick map of mean `Depth` can be created by:

```{r value-depth}
plot_datras_overview(mini,
                     metric = "mean",
                     value_var = "Depth")
```

or as gridded version with squareroot transformation:

```{r value-depth-grid}
plot_datras_overview(mini,
                     mode = "grid",
                     metric = "mean",
                     value_var = "Depth",
                     transform = "sqrt")
```

If your data set contains the number of individuals for example by using the
`DATRASextra` workflow:

```{r prep-dab}
dab <- add_total_numbers_by_haul(dab)
```

Then the plotting function can be used to generate an overview of the hauls with
the largest number of individuals:

```{r value-dab}
plot_datras_overview(dab,
                     metric = "mean",
                     value_var = "HaulN")
```

or as a gridded version:

```{r value-dab-grid}
plot_datras_overview(dab,
                     mode = "grid",
                     metric = "mean",
                     value_var = "HaulN")
```

If in addition, a meaningful offset variable is available, such as haul duration
or swept area for example by:


```{r prep-dab-sweptarea}
dab <- add_swept_area(dab)
```

Then this information can also be incorporated and the hauls with the largest
numbers of individuals per offset can be plotted:

```{r value-dab-hauldur}
plot_datras_overview(dab,
                     metric = "mean",
                     value_var = "HaulN",
                     offset_var = "HaulDur")
```

or as a gridded version with swept area:

```{r value-dab-sweptarea}
plot_datras_overview(dab,
                     mode = "grid",
                     metric = "mean",
                     value_var = "HaulN",
                     offset_var = "SweptArea")
```



## Troubleshooting

- If `spatial_basis = "statrec"` fails, check that `StatRec` exists and has
  valid values.
- If no rows appear, verify `xlim`/`ylim` filters and coordinate columns.
- If legends are suppressed in multi-panel mode, set `legend_mode = "global"`.
