---
title: "Step-by-step guide to work with the ICES' DATRAS database"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

The goal of the DATRASextra is to provide useful additional functions to work
with [DTU Aqua's *DATRAS* R package](https://github.com/DTUAqua/DATRAS) and
[ICES' DATRAS
database](https://www.ices.dk/data/data-portals/pages/datras.aspx). To get
started, make sure to install the most recent package version from GitHub.

**Outline:**

1. Download DATRAS
2. Read-in DATRAS
3. Clean DATRAS
4. Get Weight by haul
5. Summary
6. References


The package is loaded into the R environment with:

```{r}
library(DATRASextra)
```


## *Download DATRAS*


The `downloadDATRAS` function allows us to download information for any of these
surveys. If the arguments `surveys` and `years` are not specified (`NULL`), all
surveys and years are downloaded, wich can take a while. For demonstration
purposes, we download year 2023 of the NS-IBTS survey into a temporary
directory:

```{r, fig.alt = "", eval = FALSE}
survey <- "SNS"
tmp <- tempdir()
downloadDATRAS(surveys = survey, years = 2023, dir = tmp)
```

If no directory (argument `dir`) is specified, the data is downloaded into the
current working directory (`getwd()`). The function compresses downloaded data
into "zip" archives. Note, that if a year is specified in which a specific
survey did not operate, an empty file is created in the specified folder.
However, the next function just reads in non-empty files.


## *Read-in DATRAS*

Next, we can read the downloaded survey information into R:

```{r, fig.alt = "", eval = FALSE}
surv0 <- readDATRAS(file.path(tmp, survey))
```

The path can point to a specific file that you want to read in or to a whole
directory with any number of files.


## *Clean DATRAS*


Now, the `surv0` object contains the survey information in the usual `DATRASraw`
format of the *DATRAS* package. (TODO: Link to more information about this data
type or provide more information here.) It make sense to clean the data and for
example only use valid hauls. This can be done manually with the `subset()`
function or by using some recommended cleaning steps with the `clean()`
function. This function also allows us to make a relevant subset of the database
based on our needs.

```{r, fig.alt = ""}
data("dab")
surv0 <- dab
```


```{r, fig.alt = ""}
surv <- clean(surv0)
```

By default, this function also imputes missing depth information.


We can plot the resulting survey information with:

```{r, fig.alt = ""}
plot(surv)
```



## *Get Weight by haul*




## *Summary*

This vignette introduced the main features and workflow of the *DATRASextra*
package for ...



## *References*
